### 2023-02-03 products/manuscript/Manuscript.qmd
Part 1 of project
Part 1 components in manuscript.qmd, changed README file, renamed rproj

---
title: "Sara Benist MADA Project"
subtitle: "Tuberculosis Burden and Health Inequality Measures"
author: Sara Benist
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

*(Part 1 of project: data source, description, loading into raw folder, general aspects)*

The World Health Organization collected data on the inequity surrounding tuberculosis, HIV, and malaria for the [State of inequality report](https://www.who.int/data/inequality-monitor/publications/report_2021_hiv_tb_malaria), and I will be exploring the dataset for TB. More information about the data can be found [here](https://heatrepository.blob.core.windows.net/documents/data-repository-indicator-list.pdf?sp=r&st=2022-06-07T14:16:36Z&se=2023-12-30T23:16:36Z&spr=https&sv=2020-08-04&sr=b&sig=4kzThU1QDo55UOQyhWcUd8rPWJ9LxUZdRRI3zl6wKNs%3D) under "Tuberculosis Indicators". The dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators".

For the tuberculosis dataset, the data was collected from the WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources. The WHO organized the dataset to be used with the Health Equity Assessment Toolkit which is the built in data analysis and exploration tool. The database contains 10 variables regarding the burden, detection, prevention, knowledge, and social protection and observations for 194 countries over various years. The observations for each variable is further separated by up 7 inequality domains. Not all countries have data available for each year or for each inequality domain. The dataset contains a total of 7473 observations.

```{r load data}
raw_data <- readxl::read_xlsx(path = "../../data/raw_data/202206-repository-tb.xlsx")
str(raw_data)
```

## Questions/Hypotheses to be addressed

*(Part 1 of project: RQ, outcomes, predictors, patterns)*

Research question: How do the TB indicators relate to the TB incidence, prevalence, and mortality of a country, and what inequality measure shows the greatest disparity in TB outcomes?

The overall outcome I would study is TB mortality since effective health programs ideally reduce disease-specific mortality. By the end of my analysis, I would like to be able to identify populations that could be a focus of TB health improvement programs. Other outcomes I would like to explore include regional differences in drug-resistant TB and the attitudes and perceptions for high burden areas compared to medium or low burden regions.

In addition to the inequality and indicator predictors provided with the dataset, I would like to examine differences in TB outcomes based on level of TB burden. I would need to add another classification based on the literature to indicate what levels of TB incidence and prevalence would fall into each level of disease burden.

In the data, the pattern I would expect to see is a higher burden of disease on populations with greater inequality. However, I am not confident in predicting how the indicator categories, specifically the TB attitudes and perceptions, would relate to the TB outcomes.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

{{< pagebreak >}}

# Methods

*(Part 1 of project: initial analysis thoughts)*

To analyze, I would need to determine how to handle the missing data. The years between countries are not specific, and not all indicators were collected for each country. At the moment, I would subset the data based on indicator and remove incomplete observations to analyze each indicator individually. For the final analysis, I will most likely focus on TB indicators that are significantly different between subgroups or complete enough for further analysis. As the class progresses, I look forward to learning other analysis techniques. Eventually, I would like to create a dataset that can produce the statistical information, plots, and models that explore TB outcomes based on inequality measures, TB indicators, and/or level of burden.

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","height_weight_stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
Part 2 of project
data loading, cleaning, exploratory analysis, update manuscript + all readme files

---
title: "Exploratory Analysis"
author: "Sara Benist"
date: "2023-02-24"
output: html_document
---

# Processing script with R script code

This Quarto file runs the exploratory analysis code from the `exploratoryanalysis.r` script. The processed data is examined using summary tables and figures.

To start, the needed packages are loaded and the `processeddata.rda` is loaded from the `processed_data` folder.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup
The `exploratoyranalysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('exploratoryanalysis.R')
```

Next, load the needed packages.
```{r, packages, message = FALSE, warning = FALSE}
```

Then, load the data from the `processed_data` folder.
```{r, loaddata}
```

# Data exploration through tables

I would like to look at each variable individually to see an overview of the indicator. These summary tables will be saved in the `results` folder. Similar to the cleaning code, I will separate the summary tables into groups of indicators. 
```{r, exploratorytables1}
```
The BCG coverage indicator has 4 dimensions and 12 subgroups and appears to have mostly data from smaller population sizes (most of the data has less than 720) and high BCG coverage (mean of 88.7%).

The catastrophic costs has 2 dimensions and 7 subgroups, but there does not appear to be any major trends with the overview. The population is more uniform but missing data, and the percentage estimate does not have any large trends.

The case detection rate has 2 dimensions and 4 subgroups. All data is from 2020 and most of the population sizes are below 27000. The case detection rate has an average of 56.7%.

Drug resistance studies 2 subgroups within a dimension, only has data from 2019, and tends to have a low population size and percentage estimate.

```{r, exploratorytables2}
```

TB incidence studies 2 subgroups within a dimension and only has data from 2020. The population sizes are much higher than the other variables, and the incidence is a continuous variable (rather than percentage like many of the other indicators). Most observations appear to have low TB incidience.

TB mortality is similar to the incidence variable in terms of subgroups, year, and population. The mortality outcome is also continuous, and the highest mortality rate is 140 deaths/100000 population.
```{r, exploratorytables3}
```
Prevalence to notification ratio studies 2 subgroups within a dimension. There does not appear to be a trend toward shorter or longer length of time.

TB prevalence looks at 2 subgroups within a dimension, and there appears to be trend to lower amounts of cases (mean = 458) with the highest prevalence count being 1014 cases/100,000 population.

```{r, exploratorytables4}
```

Attitudes toward keeping TB infection a secret studies 2 subgroups within a dimension for the overall measure, and when split into female and male, measures 4 dimensions with 14 subgroups. All data was collected from 2011 to 2019, and each group tends to have a relatively low percentage (means of 23.3%, 26.8%, and 19.9%).
```{r, exploratorytables5}
```
Knowledge indicators of people reporting TB is spread through cough has similar dimensions and subgroups as the attitudes indicators. Data collection also occured during the same years (2011-2019). The groups all tended to have similar levels of knowledge based on the overall view (means of 66.7%, 62.3%, 68.8%).

All summary tables are saved in the `results` folder.

# Data exploration through figures

After examining the indicators, I would like to explore the differences between the subgroups using graphs and plots.

For each indicator, I will be creating a plot using `ggplot()` to study differences between subgroups. For most of these variables, I will be using a boxplot in order to compare the summary measures between subgroups. For viewing multiple dimensions and comparing subgroups within each dimension, I will be colors and grouping through `ggplot()`.

```{r, exploratoryfigures1}
```
The first plot show BCG coverage among under 1 year olds by economic status, education, place of residence, and sex. The economic status and education groups shows a clear trend of high coverage for higher income and higher education. The place of residence shows a higher coverage for urban areas, but the rural sites show greater spread of data. The coverage seems about equal between females and males. all categories show a significant number of outliers.

The second plot shows catastrophic costs due to TB by economic level and TB resistance. Higher income seems to be correlated with lower catastrophic costs, and drug resistance has a much higher correlation with costs compared to non-drug resistant TB. 
```{r, exploratoryfigures2}
```
The third plot shows case detection rate by age and sex. Higher age groups tend to have higher detection rates (although the age group is quite large).

The forth plot shows drug resistance by sex. Neither group appears to have a higher percentage of drug resistance.
```{r, exploratoryfigures3}
```
The fifth plot and sixth plot shows TB incidence and mortality rates by country, separated by sex. The countries are ordered from lowest to highest rate, and males seem to have a higher proportion of TB cases.

The seventh plot shows prevalence to notification ratio by sex. Males tend to have a longer gap between getting TB and being notified they have TB.

The eighth plot shows TB prevalence by country, showing differences between rural and urban areas. There appears to be a small trend of urban areas having higher prevalence then rural. 
```{r, exploratoryfigures4}
```
The ninth plot shows the overall attitude towards keeping a family member's TB a secret by sex. Females tend to be correlated with a higher percentage compared to males.

The tenth and eleventh graph shows the same attitude indicator for males and females while taking into account age, economic status, education, and residence. For both groups, higher age is appears to be associated with decreased percentages. There may also be a small negative correlation between education and economic status and attitude percentage.

```{r, exploratoryfigures5}
```
The twelfth plot shows TB knowledge about spreading the disease through cough by sex. Males tended to have a slightly higher proportion compared to females.

The thirteenth and fourteenth plot shows the same knowledge question for females and males, separated into the same dimensions as the attitude indicator. For both groups, increasing economic status and education showed an association with increased knowledge about TB. Urban populations also tended to show a higher percentage of knowledge compared to rural. For age groups, the 2 older groups (20-34 and 35-49) had similar proportions, but both groups had a higher percentage than the youngest (15-19). Male data appeared to be slightly less dispersed than female responses.

In the next proportion of the exploratory analysis, I would like to examine more closely the influence of outliers (and if the data needs to be separated by burden of TB) as well as start comparing indicators to TB mortality, incidence, and prevalence.

### 2023-02-24 code/processing_code/processingfile.qmd
Part 2 of project
data loading, cleaning, exploratory analysis, update manuscript + all readme files

---
title: "Data Exploring, Cleaning, and Wrangling"
author: "Sara Benist"
date: "2023-02-24"
output: html_document
---

# Processing script with code pulled in

This is Quarto file processes code chunks located in the `processingcode.r` script.

# Setup

The script below indicates which R file the Quarto should pull from.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages for data cleaning.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

Load data into `raw_data` folder and assign to `rawdata`

```{r, loaddata}
```

# Explore data

Let's look at the data using `glimpse()` and `skim()` which both present an overview of the data. Running `gg_miss_var()` will check for missing data by each variable.

```{r, exploredata1}
```
The data has the countries listed under setting, year of the data, the indicator that is being studied, and the dimension and subgroup for each observation. The estimate column holds the actual percentage or number of interest. Other variables on the data set contain formatting information.

The `skim()` function provides more information of the missing data and the numeric summaries (mean, median, percentiles, etc). The `gg_miss_var()` helps visually display the missing data.

I would also like a list of the indicator names and a count of how many countries are in the data set. 
```{r, exploredata2}
```
There are 14 indicators and 194 countries.

# Cleaning

Now the data cleaning can get started. This phase will coincide with data exploration, so more cleaning will most likely occur in the `exploratory_analysis.qmd`.

To start, I want the indicators to move to columns using `pivot_wider()` function so each indicator has a column. I will also subset the columns I would like to keep moving into the exploratory phase. The wide version of the data will be assigned to `wide_data`.

```{r, cleandata1}
```
The `wide_data` had 7473 rows and 20 columns.

## Subset by indicator
Next, I would like to create new objects for each indicator so I can make graphs/figures later on. I will use the `filter()` and `select()` functions to subset each indicator and reassign to new objects.

Here, the BCG coverage, drug resistance, and case detection rate indicators are created. 
```{r, cleandata2}
```

Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are created.

```{r, cleandata3}
```

Here, the TB incidence and TB mortality indicators are created.

```{r, cleandata4}
```

Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are created.

```{r, cleandata5}
```

Here, the TB prevalence and prevalance to notification ratio indicators are created.

```{r, cleandata6}
```

Finally, the percentage of afftected people suffering catastrophic costs due to TB indicator is created. 
```{r, cleandata7}
```

All of these indicator subsets can now be individually graphed and the trends can be compared. Most of the data seems to be from 2004-2020, and the estimates are within the bounds of the indicator (such as all the percentages are between 0 and 100).

## Missing data

While examining the code using the summary tables, a few of the indicators have missing data.

Looking over the data more closely, the missing data (marked as `NA`) is due to a country and year having data for only one subgroup in the dimension. We can drop these observations without losing important data analysis opportunities.

```{r, cleandata8}
```

I would also like to mention the odd population minimums for BCG coverage, drug resistance, and case detection rate which are 18.9, 26, and 18 respectively. These seem like odd population counts when the maximum is in the 600,000s. I believe these numbers are due to the subsetting of the indicators by dimension subgroups, so I will not be changing these numbers.

The prevalence indicator and prevalence to notification indicator both have no population data associated with the variable. I will drop the population column for both. Catastrophic cost also is missing a lot of data for the population column, but since there is some data available for a few observations, I will leave the indicator as is.

```{r, cleandata9}
```

# Save data

The data is (mostly) finished cleaning and can be saved. 
Since I have separated the indicators, we will save each object into the `processed_data` file under `processeddata.rda`

The data is saved as an rda file rather than an RDS file in order to save each item individually. This file can be loaded into the `exploratory_analysis.qmd` for the next step.

```{r, savedata}
```

### 2023-02-24 products/manuscript/Manuscript.qmd
Part 2 of project
data loading, cleaning, exploratory analysis, update manuscript + all readme files

@@
     highlight-style: github
-bibliography: ../dataanalysis_template_references.bib
+bibliography: ../sarabenist-MADA-project_references.bib
 csl: ../apa.csl
 ---
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 library(knitr)
+library(tidyverse)
+library(dplyr)
+library(ggplot2)
 ```
@@
 
-*Write a summary of your project.*
+Full summary to be added (TBA): background of topic, objective of project, quick overview of methods, main findings
 
@@
 
-*Provide enough background on your topic that others can understand the why and how of your analysis*
+TBA: background of why and how I am studying this topic
 
@@
 The dataset contains a total of 7473 observations.
-
-```{r load data}
-raw_data <- readxl::read_xlsx(path = "../../data/raw_data/202206-repository-tb.xlsx")
-str(raw_data)
-```
 
@@
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]
-
 {{< pagebreak >}}
@@
 
-## Data aquisition
+## Data aquisition, import, and cleaning
+*(Part 2 of project: data importing and cleaning)*
 
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
+The WHO dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators".
+This data was downloaded and stored into the `raw_data` folder of the repository as `rawdata`.
+Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`
+```{r load data}
+#path to data
+data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")
 
-## Data import and cleaning
+#load data and assign to rawdata
+rawdata <- readxl::read_excel(data_location)
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+str(rawdata)
+```
+Please see `processingfile.qmd` for full code describing data importing and cleaning.
+
+The `rawdata` file contains information on the country, year of data collection, the indicator they are studying, the social determinants of health they are considering, and the subgroup of each dimension.
+The dataset is mostly complete with only a few main variables missing data points
+```{r}
+#another view of the data
+skimr::skim(rawdata)
+```
+
+The indicators being studied are shown below.
+
+```{r}
+#look at what indicators are in the dataset
+unique(rawdata$indicator_name)
+```
+
+For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and assign each indicator into its own object using `filter()` and `select()` functions.
+These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup.
+An example of the subsetting is shown below.
+```{r}
+wide_data <- rawdata %>% 
+  select(c(setting, year, indicator_name,
+           indicator_abbr, dimension, subgroup,
+           estimate, population)) %>% 
+  pivot_wider(names_from = "indicator_name", values_from = "estimate")
+slice(wide_data)
+
+# BCG coverage indicator
+bcg <- wide_data %>% 
+  filter(indicator_abbr == "bcg") %>% 
+  select(c(1,2,3,4,5,6,7))
+summary(bcg)
+```
+
+Due to the `pivot_wider()` function, there was several missing observations for countries and years that had data for only some subgroups (ex: data on males but no data on females).
+Most of these observations were dropped since no important information was lost by removing these "empty" observations.
+An example is shown below.
+```{r}
+# BCG missing population & bcg data 
+# due to missing info for dimension levels
+bcg <- bcg %>% 
+  drop_na()
+```
+
+The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase.
 
@@
 
-*Explain anything related to your statistical analyses.*
+TBA: statistical analysis description
 
@@
 ## Exploratory/Descriptive analysis
+*(Part 2 of project: exploratory analysis)*
+Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+The `processedcode` was loaded into the `exploratory_analysis.qmd` file which pulls coding from the `exploratoryanalysis.r` script.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
+#Path to data.
+data_location <- here::here("data","processed_data","processeddata.rda")
+#load data
+load(data_location)
 ```
 
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+For each indicator, I produced a summary table using `skim()` to determine the number of dimensions and subgroups as well as the summary statistics to see any overall trends in the data.
+The summary tables are stored in the `results` folder.
+An example of the code is shown below.
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","height_weight_stratified.png"))
+#incidence summary and save to file location
+summary_inc <- skimr::skim(incidence)
+print(summary_inc)
+incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
+saveRDS(summary_inc, file = incsummarytable_file)
+
+#mortality summary and save to file location
+summary_mort <- skimr::skim(mortality)
+print(summary_mort)
+mortsummarytable_file <- here("results", "summary tables", "mortsummarytable.rds")
+saveRDS(summary_mort, file = mortsummarytable_file)
+```
+The population for most variables were relatively low and skewed to the right.
+The TB incidence and mortality indicators also showed a skewed distribution.
+The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.
+
+After getting an overview of the indicators, I wanted to explore the differences between subgroups and start identifying the largest disparities.
+Using `ggplot()`, I created plots of the indicators separated by dimension and subgroup.
+For most variables, these disparities were described using the `boxplot()` function as shown below.
+
+```{r}
+p13 <- tb_cough_f %>% ggplot(aes(x=dimension,
+                                 y = `People who report TB is spread through coughing - Female (%)`,
+                                 fill = subgroup)) +
+  geom_boxplot()+
+  theme(axis.text.x = element_text(angle = 90))   
+plot(p13) # plots knowledge about TB by dimension and subgroup for females
 ```
 
+The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
+```{r}
+p6 <- mortality %>% ggplot(aes(x=fct_reorder(
+  setting, `TB mortality (deaths per 100 000 population)`),
+  y = `TB mortality (deaths per 100 000 population)`,
+  color = subgroup)) +
+    geom_point()+
+  theme(axis.text.x = element_text(angle = 90))+
+  scale_fill_brewer(palette = "Spectral")+
+  xlab("setting")
+plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
+```
+
+Most indicators showed the greatest disparities due to economic status and education level.
+Further exploratory analysis should look into level of TB burden as a possible variable that should be considered, and statistical analysis in the next step should attempt to quantify the disparities between subgroups and the impact on TB outcomes.
+
+## Basic statistical analysis
+TBA: simple statistical models for associations in data
+
 ## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
+TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here
 
@@
 ## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
+TBA: summary of findings and why it matters
 
 ## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
+TBA: strengths and limitations of analysis
 
 ## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
+TBA: take away messages, citations from bibtex
 

### 2023-02-24 code/analysis_code/exploratoryanalysis.R

@@ -1,0 +1,256 @@
+###############################
+# exploratory analysis script
+#
+#this script loads the processed data and conducts exploratory analysis. 
+#Results (tables and figures) are saved in results folder.
+
+
+## ---- packages --------
+#load needed packages
+library(here) #for data loading/saving
+library(dplyr)
+library(skimr)
+library(ggplot2) #for plots/figures
+library(forcats) #for factoring
+
+
+## ---- loaddata --------
+#Path to data.
+data_location <- here::here("data","processed_data","processeddata.rda")
+#load data
+load(data_location)
+
+## ---- exploratorytables1 --------
+#bcg coverage summary and save to file location
+summary_bcg <- skimr::skim(bcg)
+print(summary_bcg)
+bcgsummarytable_file <- here("results", "summary tables", "bcgsummarytable.rds")
+saveRDS(summary_bcg, file = bcgsummarytable_file)
+
+#catastrophic cost summary and save to file location
+summary_catacost <- skimr::skim(catacost)
+print(summary_catacost)
+catacostsummarytable_file <- here("results", "summary tables", "catacostsummarytable.rds")
+saveRDS(summary_catacost, file = catacostsummarytable_file)
+
+#case detection rate summary and save to file location
+summary_cdr <- skimr::skim(cdr)
+print(summary_cdr)
+cdrsummarytable_file <- here("results", "summary tables", "cdrsummarytable.rds")
+saveRDS(summary_cdr, file = cdrsummarytable_file)
+
+#drug resistance summary and save to file location
+summary_drug <- skimr::skim(drug_resistance)
+print(summary_drug)
+drugsummarytable_file <- here("results", "summary tables", "drugsummarytable.rds")
+saveRDS(summary_drug, file = drugsummarytable_file)
+
+## ---- exploratorytables2 --------
+#incidence summary and save to file location
+summary_inc <- skimr::skim(incidence)
+print(summary_inc)
+incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
+saveRDS(summary_inc, file = incsummarytable_file)
+
+#mortality summary and save to file location
+summary_mort <- skimr::skim(mortality)
+print(summary_mort)
+mortsummarytable_file <- here("results", "summary tables", "mortsummarytable.rds")
+saveRDS(summary_mort, file = mortsummarytable_file)
+
+## ---- exploratorytables3 --------
[…]

### 2023-02-24 code/processing_code/processingcode.R

@@ -2,11 +2,9 @@
 # processing script
 #
 #this script loads the raw data, processes and cleans it 
-#and saves it as Rds file in the processed_data folder
+#and saves it as Rda file in the processed_data folder
 #
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+
 
 
 ## ---- packages --------
@@ -16,115 +14,197 @@ library(dplyr) #for data processing/cleaning
 library(tidyr) #for data processing/cleaning
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
+library(naniar) #to look for missing data
 
 ## ---- loaddata --------
 #path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","exampledata.xlsx")
+data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")
 
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
+#load data and assign to rawdata
 rawdata <- readxl::read_excel(data_location)
 
-# We can look in the Codebook (second tab) for a variable explanation
-codebook <- readxl::read_excel(data_location, sheet ="Codebook")
-print(codebook)
 
-
-## ---- exploredata --------
+## ---- exploredata1 --------
 #take a look at the data
 dplyr::glimpse(rawdata)
 
-#another way to look at the data
-summary(rawdata)
-
-#yet another way to get an idea of the data
-head(rawdata)
-
-#this is a nice way to look at data
+#another view of the data
 skimr::skim(rawdata)
 
+#check for missing data by each column
+gg_miss_var(rawdata)
 
+## ---- exploredata2 --------
+#look at what indicators are in the dataset
+unique(rawdata$indicator_name)
+
+#count of the number of countries
+length(unique(rawdata$setting))
 
 
 ## ---- cleandata1 --------
-# Inspecting the data, we find some problems that need addressing:
[…]

### 2023-03-17 code/analysis_code/exploratory_analysis.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

@@
 
-To start, the needed packages are loaded and the `processeddata.rda` is loaded from the `processed_data` folder.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
+To start, the needed packages are loaded, and the `processeddata.rda` is loaded from the `processed_data` folder.
 
 # Setup
-The `exploratoyranalysis.r` script needs to be indicated as the file from which to pull code chunks from. 
+The `exploratoryanalysis.r` script needs to be indicated as the file from which to pull code chunks from. 
 ```{r, include=FALSE, cache=FALSE}
@@
 
-For each indicator, I will be creating a plot using `ggplot()` to study differences between subgroups.
+For each indicator, I will be creating a plot using the `highburden` data set and `ggplot()` to study differences between subgroups.
 For most of these variables, I will be using a boxplot in order to compare the summary measures between subgroups.
@@
 ```
-The fifth plot and sixth plot shows TB incidence and mortality rates by country, separated by sex.
-The countries are ordered from lowest to highest rate, and males seem to have a higher proportion of TB cases.
+The 5a plot and 5b plots show TB incidence by country, separated by sex for the high burden countries and full data set.
+The countries are ordered from lowest to highest rate, and males seem to have a higher proportion of TB cases.The `highburden` data set appears to include most of the countries that have at least one new case per 100,000.
+
+The sixth plot shows TB mortality by country, separated by sex. The countries are ordered from lowest to highest rate, and males tend to have a mortality rate.
 
@@
 
-The eighth plot shows TB prevalence by country, showing differences between rural and urban areas. There appears to be a small trend of urban areas having higher prevalence then rural. 
+The 8a and 8b plots show TB prevalence by country, showing differences between rural and urban areas for the `highburden` and full data set.
+There appears to be a small trend of urban areas having higher prevalence then rural.
 ```{r, exploratoryfigures4}
@@
 
-In the next proportion of the exploratory analysis, I would like to examine more closely the influence of outliers (and if the data needs to be separated by burden of TB) as well as start comparing indicators to TB mortality, incidence, and prevalence.
+In the next proportion of the exploratory analysis, I would like to examine more closely the influence of outliers as well as start comparing indicators to TB mortality, incidence, and prevalence.

### 2023-03-17 code/analysis_code/statistical_analysis.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

---
title: "Statistical Analysis"
author: "Sara Benist"
date: "2023-03-17"
output: html_document
---

# Processing script with R script code

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The processed data is used for model fitting.

To start, the needed packages are loaded, and the `processeddata.rda` is loaded from the `processed_data` folder.

# Setup
The `statistical_analysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, cache=FALSE}
knitr::read_chunk('statistical_analysis.R')
```

# Load packages and data
The packages and data are loaded.
```{r, packages, message = FALSE, warning = FALSE}
```

```{r, loaddata}
```
# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus on linear models of the outcomes TB incidence, TB mortality, and TB prevalence. I will be fitting two models using the full data set and only the high burden countries.

## Linear model fitting for TB Incidence

```{r, fitmodel1}
```
The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Since many of the countries in the full data set have very low incidence, the coefficients for the intercept and subgroup variable are much lower. The model has more data points with low values.

### Prediction model for TB Incidence
We can also try to predict the approximate TB incidence for a group. For this example, we will include subgroup and setting as predictors.  
```{r, pred1}
```
With the prediction model including setting, there are a lot of variables in the model. The prediction would have a 1 for the country of interest and 0 for all other setting variables. For example, the prediction for a male in Central African Republic would appear as `TB incidence = 126(1)+189(1)`.

As an example, I will predict the incidence rate for males and females in India. The code for prediction modeling is based off of the `tidymodels` guidance. The plot shows the incidence rate and confidence intervals for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) in India. This can be repeated with the high burden countries.

## Linear model fitting for TB Mortality

```{r, fitmodel2}
```
For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.

## Linear model fitting for TB Prevalence

```{r, fitmodel3}
```
For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`. The difference between the high burden and full data models is lower than the other outcomes, most likely due to the long duration of TB or less data points available with the full data model.

## Linear model fitting for BCG Coverage
Now we can look at some of the other indicators in the data set. I will only be looking at the indicators that had variation between subgroups. 
```{r, fitmodel4}
```
The BCG coverage has 12 variables that produce a linear equation of `BCG coverage % = 88.4 + 0.363(Male) -10.6(no education) + 0.169(primary education) + 6.66(higher education) - 6.46(quintile 1) - 1.6(quintile 2) +1.06(quintile 3) + 4.50(quintile 4) + 7.01(quintile 5) - 2.12(rural) + 5.10(Urban)`. The full model has an equation of `BCG coverage % = 88.3 + 0.245(Male) -9.44(no education) + 0.958(primary education) + 6.42(higher education) - 6.40(quintile 1) - 1.68(quintile 2) + 0.773(quintile 3) + 3.84(quintile 4) + 6.41(quintile 5) - 2.08(rural) + 4.75(Urban)`. The two models are overall pretty similar and can be used in the same way as the prediction model (entering in 1 or 0 depending on categorization).

## Linear model fitting for Catastrophic costs

```{r, fitmodel5}
```
The high burden model has an equation of `% catastrophic costs = 87.3 - 40.4(drug-suseptible TB) - 18.6(quintile 1) - 38.8(quintile 2) - 44.6(quintile 3) - 47.4(quintile 4) - 58.5(quintile 5)`. The full model data has an equation of `% catastrophic costs = 83.5 - 35.1(drug-suseptible TB) - 11.1(quintile 1) - 30.9(quintile 2) - 38.4(quintile 3) - 41.4(quintile 4) - 54.9(quintile 5)`.
## Linear model fitting for case detection rate

```{r, fitmodel6}
```
For high burden countries, the equation is `case detection rate % = 37.9 + 19.1(15+ or older) + 16.7(female) + 16.4(male)`. The full data model has an equation of `case detection rate % = 43.8 + 18.9(15+ or older) + 15.3(female) + 16.9(male)`.

## Linear model fitting for prevalence to notification ratio

```{r, fitmodel7}
```
The high burden model has an equation of `prevalence to notification ratio in years = 1.73 + 0.664(male)`, and the full model has an equation of `prevalence to notification ratio in years = 1.73 + 0.701(male)`.

## Linear model fitting for attitudes

```{r, fitmodel8}
```
The attitude question regarding TB was about the desire to keep a diagnosis hidden. Since males are more likely to be diagnosed with TB, I focused on the male attitude and knowledge. For high burden countries, the equation is `% wanting to keep TB secret = 27.6 - 5.27(20-34 years) - 8.21(35-49 years) - 3.53(no education) - 2.71(primary education) - 7.23(higher education) - 5.01(quintile 1) - 5.23(quintile 2) - 4.97(quintile 3) - 4.38(quintile 4) - 5.86(quintile 5) - 4.91(rural) - 6.27(urban)`. For the full model, the equation is `% wanting to keep TB secret = 24.7 - 4.94(20-34 years) - 8.14(35-49 years) - 4.32(no education) - 3.08(primary education) - 6.41(higher education) - 4.44(quintile 1) - 5.24(quintile 2) - 5.18(quintile 3) - 4.59(quintile 4) - 6.20(quintile 5) - 4.66(rural) - 5.99(urban)`.

## Linear model fitting for knowledge

```{r, fitmodel9}
```
The knowledge question asked about if TB is spread through cough. We will again focus on male knowledge.

The high burden countries had an equation of `% responding TB is spread through cough = 65.9 + 4.83(20-34 years) + 6.19(35-49 years) - 9.91(no education) - 2.81(primary education) + 12.6(higher education) - 6.01(quintile 1) - 0.464(quintile 2) + 2.32(quintile 3) + 7.50(quintile 4) + 14.6(quintile 5) + 0.419(rural) + 11.8(urban)`. The full model had an equation of `% responding TB is spread through cough = 64.6 + 6.17(20-34 years) + 7.29(35-49 years) - 8.55(no education) - 1.66(primary education) + 12.8(higher education) - 3.88(quintile 1) + 0.028(quintile 2) + 3.03(quintile 3) + 8.18(quintile 4) + 15.8(quintile 5) + 1.28(rural) + 12.8(urban)`.

I would like to try and create a dataset using the high burden countries in order to predict the main outcomes of TB incidence and mortality from the indicators (BCG Coverage, knowledge, attitudes, etc). Right now, the data is not set up to complete this analysis, but I will try to find a way to do this.

### 2023-03-17 code/processing_code/processingfile.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

@@
 ```
+
 The data has the countries listed under setting, year of the data, the indicator that is being studied, and the dimension and subgroup for each observation.
@@
 I would also like a list of the indicator names and a count of how many countries are in the data set.
+
 ```{r, exploredata2}
 ```
+
 There are 14 indicators and 194 countries.
@@
 
+I will also be looking specifically at high TB burden countries.
+The high burden countries list is updated by the WHO and can be found [here](https://www.who.int/news/item/17-06-2021-who-releases-new-global-lists-of-high-burden-countries-for-tb-hiv-associated-tb-and-drug-resistant-tb).
+I will subset these countries in the `highburden` object.
+
 ```{r, cleandata1}
 ```
-The `wide_data` had 7473 rows and 20 columns.
+
+The `wide_data` had 7473 rows and 20 columns. The `highburden` object has 3761 rows and 20 columns.
 
 ## Subset by indicator
-Next, I would like to create new objects for each indicator so I can make graphs/figures later on.
-I will use the `filter()` and `select()` functions to subset each indicator and reassign to new objects.
 
-Here, the BCG coverage, drug resistance, and case detection rate indicators are created. 
+Next, I would like to check each indicator for potential data cleaning issues I need to address.
+I will use the `filter()` and `select()` functions to subset each indicator and `summary()` to check each indicator.
+
+Here, the BCG coverage, drug resistance, and case detection rate indicators are checked.
+
 ```{r, cleandata2}
@@
 
-Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are created.
+Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are checked.
 
@@
 
-Here, the TB incidence and TB mortality indicators are created.
+Here, the TB incidence and TB mortality indicators are checked.
 
@@
 
-Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are created.
+Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are checked.
 
@@
 
-Here, the TB prevalence and prevalance to notification ratio indicators are created.
+Here, the TB prevalence and prevalence to notification ratio indicators are checked.
 
@@
 
-Finally, the percentage of afftected people suffering catastrophic costs due to TB indicator is created. 
+Finally, the percentage of affected people suffering catastrophic costs due to TB indicator is checked.
+
 ```{r, cleandata7}
@@
 
-All of these indicator subsets can now be individually graphed and the trends can be compared.
 Most of the data seems to be from 2004-2020, and the estimates are within the bounds of the indicator (such as all the percentages are between 0 and 100).
@@
 Looking over the data more closely, the missing data (marked as `NA`) is due to a country and year having data for only one subgroup in the dimension.
-We can drop these observations without losing important data analysis opportunities.
-
-```{r, cleandata8}
-```
+Using the `drop_na()` function would drop all of the data since every row is incomplete.
+For now, I won't change the data until I have explored more.
 
@@
 
-The prevalence indicator and prevalence to notification indicator both have no population data associated with the variable.
-I will drop the population column for both.
-Catastrophic cost also is missing a lot of data for the population column, but since there is some data available for a few observations, I will leave the indicator as is.
-
-```{r, cleandata9}
-```
-
 # Save data
 
-The data is (mostly) finished cleaning and can be saved. 
-Since I have separated the indicators, we will save each object into the `processed_data` file under `processeddata.rda`
+The data is (mostly) finished cleaning and can be saved. Since I have separated the indicators, we will save each object into the `processed_data` file under `processeddata.rda`
 

### 2023-03-17 products/manuscript/Manuscript.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

@@
 library(ggplot2)
+library(tidymodels)
 ```
@@
 The `rawdata` file contains information on the country, year of data collection, the indicator they are studying, the social determinants of health they are considering, and the subgroup of each dimension.
-The dataset is mostly complete with only a few main variables missing data points
-```{r}
-#another view of the data
-skimr::skim(rawdata)
-```
-
+The dataset is mostly complete with only a few main variables missing data points.
 The indicators being studied are shown below.
@@
 
-For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and assign each indicator into its own object using `filter()` and `select()` functions.
+For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden`.
+I also checked each indicator using `filter()`, `select()`, and `summary()` functions.
 These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup.
-An example of the subsetting is shown below.
+An example of the data cleaning is shown below.
 ```{r}
@@
   pivot_wider(names_from = "indicator_name", values_from = "estimate")
-slice(wide_data)
+head(wide_data)
 
 # BCG coverage indicator
-bcg <- wide_data %>% 
+wide_data %>% 
   filter(indicator_abbr == "bcg") %>% 
-  select(c(1,2,3,4,5,6,7))
-summary(bcg)
+  select(c(1,2,3,4,5,6,7)) %>% 
+  summary()
 ```
@@
 Due to the `pivot_wider()` function, there was several missing observations for countries and years that had data for only some subgroups (ex: data on males but no data on females).
-Most of these observations were dropped since no important information was lost by removing these "empty" observations.
-An example is shown below.
+Since I will be maintaining all the indicators in their own data set, `drop_na()` would remove all data points.
+Instead, I will handle missing data points after exploring the data further.
+
+The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase. I have loaded the processed data below.
 ```{r}
-# BCG missing population & bcg data 
-# due to missing info for dimension levels
-bcg <- bcg %>% 
-  drop_na()
+#path to data
+data_location <- here::here("data","processed_data","processeddata.rda")
+#load data. 
+load(data_location)
 ```
 
-The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase.
+## Statistical analysis
+*(Part 3 of project: statistical analysis)*
+For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator.
+An example is shown below.
+This method was repeated for the `wide_data` and `highburden` countries.
 
-## Statistical analysis
+```{r}
+# fit linear model using TB incidence as outcome, subgroup as predictor
+lm_mod <- linear_reg() #set linear regression
 
-TBA: statistical analysis description
+#fit model with high burden countries
+inclm_fit <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
+      data = highburden) #fit linear model
+incfittable <- tidy(inclm_fit) 
+print(incfittable) #produce tidy table of fitted model
+```
+
+Other indicators were also analyzed and fit to a linear model as shown below. 
+```{r}
+# fit linear model with BCG coverage as outcome, subgroups as predictors
+
+#fit model with high burden countries
+BCGlm_fit <- lm_mod %>% 
+  fit(`BCG immunization coverage among one-year-olds (%)` ~ subgroup,
+      data = highburden) #fit linear model
+BCGfittable <- tidy(BCGlm_fit) 
+print(BCGfittable) #produce tidy table of fitted model
+```
+
+In the next step of my project, I would like to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes.
+This will most likely be limited to the high burden countries.
 
@@
 #incidence summary and save to file location
-summary_inc <- skimr::skim(incidence)
+summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
 print(summary_inc)
-incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
-saveRDS(summary_inc, file = incsummarytable_file)
 
 #mortality summary and save to file location
-summary_mort <- skimr::skim(mortality)
+summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
 print(summary_mort)
-mortsummarytable_file <- here("results", "summary tables", "mortsummarytable.rds")
-saveRDS(summary_mort, file = mortsummarytable_file)
 ```
@@
 ```{r}
-p13 <- tb_cough_f %>% ggplot(aes(x=dimension,
+p13 <- highburden %>%
+  filter(!dimension %in% c("Age (2 groups) (0-15+)", "TB drug resistance", "Sex")) %>%
+  ggplot(aes(x=dimension,
              y = `People who report TB is spread through coughing - Female (%)`,
@@
 ```{r}
-p6 <- mortality %>% ggplot(aes(x=fct_reorder(
+p6 <- highburden %>% 
+  filter(dimension %in% "Sex") %>%
+  filter(indicator_abbr == "mortality") %>% 
+  select(c(1,2,3,4,5,6,13)) %>% 
+  ggplot(aes(x=fct_reorder(
   setting, `TB mortality (deaths per 100 000 population)`),
@@
 ## Basic statistical analysis
+*(Part 3 of project: linear model fitting and predictions, more to be added)*
 TBA: simple statistical models for associations in data
 
+The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
+The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`.
+Based on the p-value for the subgroup variable, sex has a significant association with TB incidence in both models.
+Males having higher TB incidence is congruent with the exploratory analysis findings.
+The full model has smaller coefficients compared to `highburden` model.
+```{r}
+readRDS(file = here("results", "model fit tables", "incfittable.rds")) #TB incidence model for high burden countries
+readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) #TB incidence model for full data
+```
+The fitted model summaries below are the TB mortality for high burden countries and full data.
+For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
+Sex is a statistically significant predictor of TB mortality (p-values <0.05), but this does not consider any other factors.
+The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators.
+```{r}
+readRDS(file = here("results", "model fit tables", "mortfittable.rds")) #TB mortality model for high burden countries
+readRDS(file = here("results", "model fit tables", "mortfittableFD.rds")) #TB mortality model for full data
+```
+The last outcome model I fitted was TB prevalence for high burden countries and the full data.
+For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0.
+The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.
+```{r}
+readRDS(file = here("results", "model fit tables", "prevfittable.rds")) #TB prevalence model for high burden countries
+readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) #TB prevalence model for full data
+```
+The models can be further explored by including `setting` as a predictor, shown in the prediction model below.
+The prediction model is basic prediction of TB incidence from subgroup and setting.
+As an example, I predicted the incidence for females and males in India (shown in the plot).
+The estimates for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) are shown using `geom_point()` and `geom_errorbar()`.
+I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables.
+```{r}
+#prediction model for TB incidence with subgroup and setting as predictors
+predinclm_fit <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
+      data = highburden) #fit linear model
+predincfittable <- tidy(predinclm_fit) 
+print(predincfittable)
+
+#prediction plot using the above model estimating TB incidence for females and males in India
+readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
+```
 ## Full analysis

### 2023-03-17 code/analysis_code/exploratoryanalysis.R

@@ -22,89 +22,89 @@ load(data_location)
 
 ## ---- exploratorytables1 --------
 #bcg coverage summary and save to file location
-summary_bcg <- skimr::skim(bcg)
+summary_bcg <- skimr::skim(wide_data$`BCG immunization coverage among one-year-olds (%)`)
 print(summary_bcg)
 bcgsummarytable_file <- here("results", "summary tables", "bcgsummarytable.rds")
 saveRDS(summary_bcg, file = bcgsummarytable_file)
 
 #catastrophic cost summary and save to file location
-summary_catacost <- skimr::skim(catacost)
+summary_catacost <- skimr::skim(wide_data$`Families affected by TB facing catastrophic costs due to TB (%)`)
 print(summary_catacost)
 catacostsummarytable_file <- here("results", "summary tables", "catacostsummarytable.rds")
 saveRDS(summary_catacost, file = catacostsummarytable_file)
 
 #case detection rate summary and save to file location
-summary_cdr <- skimr::skim(cdr)
+summary_cdr <- skimr::skim(wide_data$`Case detection rate (%)`)
 print(summary_cdr)
 cdrsummarytable_file <- here("results", "summary tables", "cdrsummarytable.rds")
 saveRDS(summary_cdr, file = cdrsummarytable_file)
 
 #drug resistance summary and save to file location
-summary_drug <- skimr::skim(drug_resistance)
+summary_drug <- skimr::skim(wide_data$`People with MDR/RR-TB (%)`)
 print(summary_drug)
 drugsummarytable_file <- here("results", "summary tables", "drugsummarytable.rds")
 saveRDS(summary_drug, file = drugsummarytable_file)
 
 ## ---- exploratorytables2 --------
 #incidence summary and save to file location
-summary_inc <- skimr::skim(incidence)
+summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
 print(summary_inc)
 incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
 saveRDS(summary_inc, file = incsummarytable_file)
 
 #mortality summary and save to file location
-summary_mort <- skimr::skim(mortality)
+summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
 print(summary_mort)
 mortsummarytable_file <- here("results", "summary tables", "mortsummarytable.rds")
[…]

### 2023-03-17 code/analysis_code/statistical_analysis.R

@@ -4,53 +4,259 @@
 #this script loads the processed, cleaned data, does a simple analysis
 #and saves the results to the results folder
 
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
+## ---- packages --------
+#load needed packages
+library(here) 
+library(tidymodels)
+library(tidyverse)
+library(ggplot2)
 
+
+## ---- loaddata --------
 #path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","processeddata.rds")
+data_location <- here::here("data","processed_data","processeddata.rda")
 
 #load data. 
-mydata <- readRDS(data_location)
+load(data_location)
 
-
+## ---- fitmodel1 ---------
 ######################################
 #Data fitting/statistical analysis
 ######################################
 
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
 
-lmfit1 <- lm(Height ~ Weight, mydata)  
+# fit linear model using TB incidence as outcome, subgroup as predictor
+lm_mod <- linear_reg() #set linear regression
 
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
+#fit model with high burden countries
+inclm_fit <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
+      data = highburden) #fit linear model
+incfittable <- tidy(inclm_fit) 
+print(incfittable) #produce tidy table of fitted model
 
-#look at fit results
-print(lmtable1)
+table_file1 = here("results", "model fit tables", "incfittable.rds")
+saveRDS(incfittable, file = table_file1) #save summary table
 
-# save fit results table  
-table_file1 = here("results", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
+#fit model with full data
+inclm_fitFD <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
+      data = wide_data) #fit linear model
+incfittable2 <- tidy(inclm_fitFD) 
+print(incfittable2) #produce tidy table of fitted model
 
-############################
[…]

### 2023-03-17 code/processing_code/processingcode.R

@@ -44,154 +44,126 @@ length(unique(rawdata$setting))
 
 ## ---- cleandata1 --------
 # Pivot data wider so each indicator has a column
-wide_data <- rawdata %>% 
+ wide_data <- rawdata %>% 
   select(c(setting, year, indicator_name,
            indicator_abbr, dimension, subgroup,
            estimate, population)) %>% 
   pivot_wider(names_from = "indicator_name", values_from = "estimate")
 slice(wide_data)
 
+#subset high burden countries
+highburden <- wide_data %>% 
+  filter(setting %in% c("Brazil", "Central African Republic", "Congo",
+           "Ethiopia", "Gabon", "Kenya",
+           "Lesotho", "Liberia", "Namibia",
+           "Thailand", "Uganda", "United Republic of Tanzania",
+           "Botswana", "Cameroon", "Eswatini",
+           "Guinea", "Guinea-Bissau", "Malawi",
+           "Russian Federation", "Zimbabwe", "China",
+           "Democratic Republic of the Congo", "India",
+           "Indonesia", "Mozambique", "Myanmar",
+           "Nigeria", "Philippines", "South Africa",
+           "Zambia", "Sierra Leone", "Angola",
+           "Bangladesh", "Democratic People's Republic of Korea",
+           "Mongolia", "Pakistan", "Papua New Guinea",
+           "Viet Nam", "Azerbaijan", "Belarus", "Kazakhstan",
+           "Kyrgyzstan", "Nepal", "Peru", "Republic of Moldova",
+           "Somalia", "Tajikistan", "Ukraine",
+           "Uzbekistan", "Zimbabwe")) 
+unique(highburden$setting)
 
 ## ---- cleandata2 --------
-# create new subsets by indicator since I want to compare 
-# each indicator to TB outcomes
+# look at each indicator to check for cleaning requirements
 unique(wide_data$indicator_abbr)
 
 # BCG coverage indicator
-bcg <- wide_data %>% 
+wide_data %>% 
   filter(indicator_abbr == "bcg") %>% 
-  select(c(1,2,3,4,5,6,7))
-summary(bcg)
+  select(c(1,2,3,4,5,6,7)) %>% 
+  summary()
 
[…]

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
Part 4a of project
machine learning analysis

@@
 author: "Sara Benist"
-date: "2023-03-17"
+date: "2023-04-06"
 output: html_document
@@
 
-This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The processed data is used for model fitting.
+This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script.
+The best way to view the output of the script is to render the qmd and view the html document.
+The processed data is used for model fitting.
 
@@
 Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set.
-First, we will focus on linear models of the outcomes TB incidence, TB mortality, and TB prevalence.
-I will be fitting two models using the full data set and only the high burden countries.
+First, we will focus split the data into testing and training sets and set up the cross-validation folds.
+The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.
 
-## Linear model fitting for TB Incidence
+```{r, splitdata}
 
-```{r, fitmodel1}
 ```
-The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
-The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`.
-Since many of the countries in the full data set have very low incidence, the coefficients for the intercept and subgroup variable are much lower.
-The model has more data points with low values.
 
-### Prediction model for TB Incidence
-We can also try to predict the approximate TB incidence for a group. For this example, we will include subgroup and setting as predictors.  
-```{r, pred1}
+I will be tuning decision tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the full data set.
+
+## Decision tree model for TB Incidence
+
+```{r, wdincidencemodel}
 ```
-With the prediction model including setting, there are a lot of variables in the model.
-The prediction would have a 1 for the country of interest and 0 for all other setting variables.
-For example, the prediction for a male in Central African Republic would appear as `TB incidence = 126(1)+189(1)`.
+The best tree model for TB incidence using the full data has a depth of 4 with an RMSE of 128.
+This is a very low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 62.8, but using the testing data, the model had a performance of RMSE = 110.
+The most important predictor was country.
 
-As an example, I will predict the incidence rate for males and females in India.
-The code for prediction modeling is based off of the `tidymodels` guidance.
-The plot shows the incidence rate and confidence intervals for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) in India.
-This can be repeated with the high burden countries.
+```{r, hbincidencemodel}
 
-## Linear model fitting for TB Mortality
+```
+The best tree model for TB incidence using the high burden countries has a depth of 4 with an RMSE of 162.
+This is also a low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 126 which was much higher compared to the full dataset.
+Using the testing data, the model had a performance of RMSE = 134.
+The most important predictor was also country.
+## Decision tree model for TB Mortality
 
-```{r, fitmodel2}
+```{r}
+#recipe predicting TB mortality
+mortrecipe <- recipe(`TB mortality (deaths per 100 000 population)` ~.,
+                 data = wdtrain_data)
+#create tree model
+treemodel <- decision_tree(
+  cost_complexity = tune(),
+  tree_depth = tune(),
+) %>% 
+  set_engine("rpart") %>% 
+  set_mode("regression")
+
+#create workflow
+mortwflow <- workflow() %>% 
+  add_model(treemodel) %>% 
+  add_recipe(mortrecipe)
+
+#set seed for reproducibility
+set.seed(626)
+#resample using cross-validation folds to tune parameters
+morttree_resamp <- mortwflow %>% 
+  tune_grid(
+    resamples = wdfolds,
+    grid = tree_grid,
+    control = control_grid(save_pred = TRUE)
+  )
+
+#plot resampling by parameters
+morttreeplot <- morttree_resamp %>% autoplot()
+morttreeplot
+
+#show the best model
+morttree_resamp %>% 
+  show_best()
+
+#select best model
+wdmortbest_tree <- morttree_resamp %>% select_best("rmse")
+
+#create final workflow
+wdmortfinal_wf <- mortwflow %>% finalize_workflow(wdmortbest_tree)
+
+#fit the final workflow to the training data
+wdmortfinal_fit <- wdmortfinal_wf %>% fit(wdtrain_data)
+
+#find RMSE for model with training data
+wdmortfitted <- augment(wdmortfinal_fit, wdtrain_data) %>% 
+  select(`TB mortality (deaths per 100 000 population)`, .pred) %>% 
+  rmse(truth = `TB mortality (deaths per 100 000 population)`, .pred)
+wdmortfitted
+#fit final workflow to test data
+wdmortfittest <- wdmortfinal_fit %>% last_fit(wddatasplit)
+
+#find RMSE for model with test data
+wdmortfittest %>% collect_metrics()
+
+#check important variables of model
+wdmortfittest %>% extract_fit_parsnip() %>% vip()
+
 ```
-For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
+The best tree model for TB mortality using the full data has a depth of 1 with an RMSE of 16.9.
+The fitted model to the training data has a RMSE of 11.3, but using the testing data, the model had a performance of RMSE = 17.4.
+The most important predictor was country.
 
-## Linear model fitting for TB Prevalence
+```{r, hbmortalitymodel}
+```
+The best tree model for TB mortality using the high burden countries has a depth of 1 with an RMSE of 20.8.
+The fitted model to the training data has a RMSE of 16.0.
+Using the testing data, the model had a performance of RMSE = 21.0.
+The most important predictor was country.
 
-```{r, fitmodel3}
+## Decision tree model for TB Prevalence
+
+```{r, wdprevalencemodel}
 ```
-For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0.
-The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.
-The difference between the high burden and full data models is lower than the other outcomes, most likely due to the long duration of TB or less data points available with the full data model.
+The best tree model for TB prevalence using the full data has a depth of 1 with an RMSE of 208.
+This is a very low level of performance in predicting TB prevalence.
+The fitted model to the training data has a RMSE of 135, and using the testing data, the model had a performance of RMSE = 287.
+The most important predictor was country.
 
-## Linear model fitting for BCG Coverage
-Now we can look at some of the other indicators in the data set. I will only be looking at the indicators that had variation between subgroups. 
-```{r, fitmodel4}
+```{r, hbprevalencemodel}
 ```
-The BCG coverage has 12 variables that produce a linear equation of `BCG coverage % = 88.4 + 0.363(Male) -10.6(no education) + 0.169(primary education) + 6.66(higher education) - 6.46(quintile 1) - 1.6(quintile 2) +1.06(quintile 3) + 4.50(quintile 4) + 7.01(quintile 5) - 2.12(rural) + 5.10(Urban)`.
-The full model has an equation of `BCG coverage % = 88.3 + 0.245(Male) -9.44(no education) + 0.958(primary education) + 6.42(higher education) - 6.40(quintile 1) - 1.68(quintile 2) + 0.773(quintile 3) + 3.84(quintile 4) + 6.41(quintile 5) - 2.08(rural) + 4.75(Urban)`.
-The two models are overall pretty similar and can be used in the same way as the prediction model (entering in 1 or 0 depending on categorization).
+The best tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 172.
+This is a very low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 138, but using the testing data, the model had a performance of RMSE = 237.
+The most important predictor was country.
 
-## Linear model fitting for Catastrophic costs
+The models were overall not a good predictor of TB outcomes.
+This is possibly due to the large variance in the outcomes as well as the low number of predictors available.
+Since country is consistently the most important predictor, we can look at some of the health indicators to see some of the underlying trends that can occur within a country.
 
-```{r, fitmodel5}
+## Decision tree model for BCG coverage
+
+```{r, wdbcgmodel}
 ```
-The high burden model has an equation of `% catastrophic costs = 87.3 - 40.4(drug-suseptible TB) - 18.6(quintile 1) - 38.8(quintile 2) - 44.6(quintile 3) - 47.4(quintile 4) - 58.5(quintile 5)`.
-The full model data has an equation of `% catastrophic costs = 83.5 - 35.1(drug-suseptible TB) - 11.1(quintile 1) - 30.9(quintile 2) - 38.4(quintile 3) - 41.4(quintile 4) - 54.9(quintile 5)`.
-## Linear model fitting for case detection rate
+The best tree model for BCG coverage using the full data has a depth of 15 with an RMSE of 5.86.
+This would appear to be a better prediction than the previous ones.
+The fitted model to the training data has a RMSE of 4.29, and using the testing data, the model had a performance of RMSE = 5.30.
+The most important predictor was country followed by subgroup.
 
-```{r, fitmodel6}
+```{r, hbbcgmodel}
 ```
-For high burden countries, the equation is `case detection rate % = 37.9 + 19.1(15+ or older) + 16.7(female) + 16.4(male)`.
-The full data model has an equation of `case detection rate % = 43.8 + 18.9(15+ or older) + 15.3(female) + 16.9(male)`.
+The best tree model for BCG coverage using high burden countries has a depth of 11 with an RMSE of 5.79.
+The fitted model to the training data has a RMSE of 4.00, but using the testing data, the model had a performance of RMSE = 5.56.
+These are approximately similar to the full data set.
+The most important predictor was country.
 
-## Linear model fitting for prevalence to notification ratio
+## Decision tree model for Catastrophic Cost
+```{r, wdcatacostmodel}
+```
+The best tree model for catastrophic costs using the full data has a depth of 4 with an RMSE of 17.2.
+The fitted model to the training data has a RMSE of 10.2, but using the testing data, the model had a performance of RMSE = 18.1.
+The most important predictor was subgroup.
 
-```{r, fitmodel7}
+```{r, hbcatacostmodel}
 ```
-The high burden model has an equation of `prevalence to notification ratio in years = 1.73 + 0.664(male)`, and the full model has an equation of `prevalence to notification ratio in years = 1.73 + 0.701(male)`.
+The best tree model for catastrophic costs using the high burden countries has a depth of 4 with an RMSE of 16.3.
+The fitted model to the training data has a RMSE of 10.7, but using the testing data, the model had a performance of RMSE = 15.2.
+The most important predictor was subgroup.
 
-## Linear model fitting for attitudes
+## Decision tree model for Case Detection Rate
 
-```{r, fitmodel8}
+```{r, wdcasedetectmodel}
 ```
-The attitude question regarding TB was about the desire to keep a diagnosis hidden.
-Since males are more likely to be diagnosed with TB, I focused on the male attitude and knowledge.
-For high burden countries, the equation is `% wanting to keep TB secret = 27.6 - 5.27(20-34 years) - 8.21(35-49 years) - 3.53(no education) - 2.71(primary education) - 7.23(higher education) - 5.01(quintile 1) - 5.23(quintile 2) - 4.97(quintile 3) - 4.38(quintile 4) - 5.86(quintile 5) - 4.91(rural) - 6.27(urban)`.
-For the full model, the equation is `% wanting to keep TB secret = 24.7 - 4.94(20-34 years) - 8.14(35-49 years) - 4.32(no education) - 3.08(primary education) - 6.41(higher education) - 4.44(quintile 1) - 5.24(quintile 2) - 5.18(quintile 3) - 4.59(quintile 4) - 6.20(quintile 5) - 4.66(rural) - 5.99(urban)`.
+The best tree model for case detection rate using the full data has a depth of 8 with an RMSE of 13.6.
+The fitted model to the training data has a RMSE of ###, but using the testing data, the model had a performance of RMSE = 12.2.
+The most important predictor was country.
 
-## Linear model fitting for knowledge
+```{r, hbcasedetectmodel}
+```
+The best tree model for case detection rate using high burden countries has a depth of 8 with an RMSE of 16.1.
+This is a very low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 7.93, but using the testing data, the model had a performance of RMSE = 11.8.
+The most important predictor was country.
 
-```{r, fitmodel9}
+## Decision tree model for Prevalence to Notification ratio
+
+```{r, wdptnmodel}
 ```
-The knowledge question asked about if TB is spread through cough. We will again focus on male knowledge.
+The best tree model for prevalence to notification ratio using the full data has a depth of 4 with an RMSE of 1.06.
+This model has the highest performance so far.
+The fitted model to the training data has a RMSE of 0.546, and using the testing data, the model had a performance of RMSE = 0.656.
+The most important predictor was country.
 
-The high burden countries had an equation of `% responding TB is spread through cough = 65.9 + 4.83(20-34 years) + 6.19(35-49 years) - 9.91(no education) - 2.81(primary education) + 12.6(higher education) - 6.01(quintile 1) - 0.464(quintile 2) + 2.32(quintile 3) + 7.50(quintile 4) + 14.6(quintile 5) + 0.419(rural) + 11.8(urban)`.
-The full model had an equation of `% responding TB is spread through cough = 64.6 + 6.17(20-34 years) + 7.29(35-49 years) - 8.55(no education) - 1.66(primary education) + 12.8(higher education) - 3.88(quintile 1) + 0.028(quintile 2) + 3.03(quintile 3) + 8.18(quintile 4) + 15.8(quintile 5) + 1.28(rural) + 12.8(urban)`.
+```{r, hbptnmodel}
+```
+The best tree model for prevalence to notification ratio using high burden countries has a depth of 1 with an RMSE of 0.984.
+This model also has high performance and has a different tree depth than the full data set.
+The fitted model to the training data has a RMSE of 0.650, but using the testing data, the model had a performance of RMSE = 0.773.
+The most important and only predictor was country.
 
-I would like to try and create a dataset using the high burden countries in order to predict the main outcomes of TB incidence and mortality from the indicators (BCG Coverage, knowledge, attitudes, etc).
-Right now, the data is not set up to complete this analysis, but I will try to find a way to do this.
+## Decision tree model for Male Attitudes
+
+```{r, wdmaleattitudemodel}
+```
+The best tree model for male attitudes using the full data has a depth of 4 with an RMSE of 5.39.
+The fitted model to the training data has a RMSE of 3.87, and using the testing data, the model had a performance of RMSE = 4.62.
+The most important predictor was country.
+
+```{r, hbmaleattitubemodel}
+```
+The best tree model for male attitude using high burden countries has a depth of 8 with an RMSE of 5.70.
+The fitted model to the training data has a RMSE of 3.71, but using the testing data, the model had a performance of RMSE = 4.46.
+The most important predictor was country.
+
+## Decision tree model for Male Knowledge
+```{r, wdmaleknowledgemodel}
+```
+The best tree model for male knowledge using the full data has a depth of 8 with an RMSE of 8.52.
+The fitted model to the training data has a RMSE of 5.69, but using the testing data, the model had a performance of RMSE = 6.49.
+The most important predictor was country.
+
+```{r, hbmaleknowledgemodel}
+```
+The best tree model for male knowledge using the high burden countries has a depth of 4 with an RMSE of 10.8.
+The fitted model to the training data has a RMSE of 6.23, but using the testing data, the model had a performance of RMSE = 8.99.
+The most important predictor was year which was different from the full data set.

### 2023-04-07 products/manuscript/Manuscript.qmd
Part 4a of project
machine learning analysis

@@
 
-TBA: background of why and how I am studying this topic
+Tuberculosis is an infectious disease caused by Mycobacterium tuberculosis and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022].
+Approximately 10 million people are infected each year, and 1.5 million die from the disease [@WHO2022].
+Humanity has been impacted by TB for centuries, but the burden of disease is not equal across countries and populations [@WHO2022b].
+The World Health Organization marked 30 countries as having highest burden of TB due to the number of cases, presence of multi-drug resistance strains, and the high mortality within certain populations such as HIV [@WHO2021b].
+To understand more about why some countries or groups of people experience the higher burden, the World Health Organization gathered data on several health equity indicators subgroups.
+This project will explore and analyze the disparities within the indicators and the effect on tuberculosis outcomes.
 
@@
 
-*(Part 1 of project: data source, description, loading into raw folder, general aspects)*
+The World Health Organization collected data on the inequity surrounding tuberculosis, HIV, and malaria for the [State of inequality report](https://www.who.int/data/inequality-monitor/publications/report_2021_hiv_tb_malaria), and I will be exploring the dataset for TB [@WHO2021].
+More information about the data can be found [here](https://heatrepository.blob.core.windows.net/documents/data-repository-indicator-list.pdf?sp=r&st=2022-06-07T14:16:36Z&se=2023-12-30T23:16:36Z&spr=https&sv=2020-08-04&sr=b&sig=4kzThU1QDo55UOQyhWcUd8rPWJ9LxUZdRRI3zl6wKNs%3D) under "Tuberculosis Indicators" [@WHOTBData].
+The dataset can be found [here](https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators" [@WHOTBData].
 
-The World Health Organization collected data on the inequity surrounding tuberculosis, HIV, and malaria for the [State of inequality report](https://www.who.int/data/inequality-monitor/publications/report_2021_hiv_tb_malaria), and I will be exploring the dataset for TB.
-More information about the data can be found [here](https://heatrepository.blob.core.windows.net/documents/data-repository-indicator-list.pdf?sp=r&st=2022-06-07T14:16:36Z&se=2023-12-30T23:16:36Z&spr=https&sv=2020-08-04&sr=b&sig=4kzThU1QDo55UOQyhWcUd8rPWJ9LxUZdRRI3zl6wKNs%3D) under "Tuberculosis Indicators".
-The dataset can be found [here](o%20https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00) under "Tuberculosis Indicators".
-
-For the tuberculosis dataset, the data was collected from the WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources.
+For the tuberculosis dataset, the data was collected from the WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources [@WHO2021].
 The WHO organized the dataset to be used with the Health Equity Assessment Toolkit which is the built in data analysis and exploration tool.
@@
 
-*(Part 1 of project: RQ, outcomes, predictors, patterns)*
-
-Research question: How do the TB indicators relate to the TB incidence, prevalence, and mortality of a country, and what inequality measure shows the greatest disparity in TB outcomes?
+Research question: Does high disparities between inequality measures show high correlation to TB incidence, prevalence, and mortality of a country?
+How do these relationships differ for high burden countries compared to all countries?
 
@@
 
-*(Part 1 of project: initial analysis thoughts)*
-
 To analyze, I would need to determine how to handle the missing data.
@@
 ## Data aquisition, import, and cleaning
-*(Part 2 of project: data importing and cleaning)*
 
@@
 Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`
+
 ```{r load data}
@@
 ```
+
 Please see `processingfile.qmd` for full code describing data importing and cleaning.
@@
 An example of the data cleaning is shown below.
+
 ```{r}
@@
 The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase. I have loaded the processed data below.
+
 ```{r}
@@
 ## Statistical analysis
-*(Part 3 of project: statistical analysis)*
+
 For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator.
@@
 Other indicators were also analyzed and fit to a linear model as shown below.
+
 ```{r}
@@
 ## Exploratory/Descriptive analysis
-*(Part 2 of project: exploratory analysis)*
-Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
+
+*(Part 2 of project: exploratory analysis)* Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
 
@@
 ```
+
 The population for most variables were relatively low and skewed to the right.
@@
 The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
+
 ```{r}
@@
 ## Basic statistical analysis
-*(Part 3 of project: linear model fitting and predictions, more to be added)*
-TBA: simple statistical models for associations in data
+
+*(Part 3 of project: linear model fitting and predictions, more to be added)* TBA: simple statistical models for associations in data
 
@@
 The full model has smaller coefficients compared to `highburden` model.
+
 ```{r}
@@
 ```
+
 The fitted model summaries below are the TB mortality for high burden countries and full data.
 For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
-Sex is a statistically significant predictor of TB mortality (p-values <0.05), but this does not consider any other factors.
+Sex is a statistically significant predictor of TB mortality (p-values \<0.05), but this does not consider any other factors.
 The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators.
+
 ```{r}
@@
 ```
+
 The last outcome model I fitted was TB prevalence for high burden countries and the full data.
@@
 The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.
+
 ```{r}
@@
 ```
+
 The models can be further explored by including `setting` as a predictor, shown in the prediction model below.
@@
 I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables.
+
 ```{r}
@@
 ```
+
 ## Full analysis
+
 TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here
@@
 ## Summary and Interpretation
+
 TBA: summary of findings and why it matters
@@
 ## Strengths and Limitations
+
 TBA: strengths and limitations of analysis
@@
 ## Conclusions
+
 TBA: take away messages, citations from bibtex

### 2023-04-07 code/analysis_code/statistical_analysis.R

@@ -10,6 +10,8 @@ library(here)
 library(tidymodels)
 library(tidyverse)
 library(ggplot2)
+library(rpart.plot)
+library(vip)
 
 
 ## ---- loaddata --------
@@ -19,244 +21,1030 @@ data_location <- here::here("data","processed_data","processeddata.rda")
 #load data. 
 load(data_location)
 
-## ---- fitmodel1 ---------
-######################################
-#Data fitting/statistical analysis
-######################################
+## ---- splitdata --------
+#split data into training and test sets for wide_data
+wddatasplit <- initial_split(wide_data, prop = 3/4)
+wdtest_data <- testing(wddatasplit)
+wdtrain_data <- training(wddatasplit)
+
+#set seed for reproducibility
+set.seed(626)
+
+#create cross-validation folds
+wdfolds <- vfold_cv(wdtrain_data, v = 5)
+
+#split data into training and test sets for highburden
+hbdatasplit <- initial_split(highburden, prop = 3/4)
+hbtest_data <- testing(hbdatasplit)
+hbtrain_data <- training(hbdatasplit)
+
+#set seed for reproducibility
+set.seed(626)
+
+#create cross-validation folds
+hbfolds <- vfold_cv(hbtrain_data, v = 5)
 
 
-# fit linear model using TB incidence as outcome, subgroup as predictor
-lm_mod <- linear_reg() #set linear regression
+## ---- wdincidencemodel ---------
+#recipe predicting TB incidence
+increcipe <- recipe(`TB incidence (new infections per 100 000 population)` ~.,
+                 data = wdtrain_data)
+#create tree model
+treemodel <- decision_tree(
+  cost_complexity = tune(),
+  tree_depth = tune(),
+) %>% 
+  set_engine("rpart") %>% 
+  set_mode("regression")
 
-#fit model with high burden countries
-inclm_fit <- lm_mod %>% 
-  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
-      data = highburden) #fit linear model
-incfittable <- tidy(inclm_fit) 
-print(incfittable) #produce tidy table of fitted model
+#create work flow
+incwflow <- workflow() %>% 
+  add_model(treemodel) %>% 
+  add_recipe(increcipe)
 
-table_file1 = here("results", "model fit tables", "incfittable.rds")
-saveRDS(incfittable, file = table_file1) #save summary table
+#create grid of tuning parameters
[…]

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
Part 4 of project
updated references, machine learning stat analysis, inserted figures into manuscript

@@
 ```
+
+# Basic modelling
+For some basic modeling, I will be fitting linear models to the TB outcomes with the subgroups as predictors.
+This will offer another view of the data.
+The prediction model also offers a way to explore the data by submitting a sample prediction of TB incidence for females and males in India.
+```{r, basicmodels}
+```
+The biggest issue with the linear modeling is the assumption that the relationship between predictors and outcomes is linear, and the model cannot be tuned and replicated.
+
 # Model fitting

### 2023-04-07 products/manuscript/Manuscript.qmd
Part 4 of project
updated references, machine learning stat analysis, inserted figures into manuscript

@@
 
+# Questions for Dr. Handel
+I would like to be able to predict the main TB outcomes based on the health equity indicators (ex: level of TB incidence predicted from % of people with drug-resistant strains), but I cannot join the rows together correctly.
+When I tried bind_rows() on the one of the countries to see if it would work, the outcome rows are still separated and if I remove year to force everything onto the same row, I have 90 variables and they are grouped within the same cell.
+I added an example below.
+Do you have any advice?
+
+```{r}
+#path to data
+data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")
+
+#load data and assign to rawdata
+rawdata <- readxl::read_excel(data_location)
+
+#without year removed: outcomes remain on a separate row
+test <- rawdata %>% 
+  select(c(setting, year, indicator_name,
+           dimension, subgroup,
+           estimate)) %>% 
+  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
+  filter(setting == "Afghanistan") %>% bind_rows()
+test
+
+#with year removed: cells contain multiple observations
+test2 <- rawdata %>% 
+  select(c(setting, indicator_name,
+           dimension, subgroup,
+           estimate)) %>% 
+  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
+  filter(setting == "Afghanistan") %>% bind_rows()
+test2
+```
+If it is not feasible, I will be analyzing TB outcomes separately from indicators instead.
+
 # Summary/Abstract
@@
 
-Tuberculosis is an infectious disease caused by Mycobacterium tuberculosis and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022].
+Tuberculosis is an infectious disease caused by *Mycobacterium tuberculosis* and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022].
 Approximately 10 million people are infected each year, and 1.5 million die from the disease [@WHO2022].
@@
 To understand more about why some countries or groups of people experience the higher burden, the World Health Organization gathered data on several health equity indicators subgroups.
-This project will explore and analyze the disparities within the indicators and the effect on tuberculosis outcomes.
+The Global Tuberculosis Programme houses the End TB Strategy with a goal to eradicating TB.
+Their current objective is to reduce TB incidence by 80% and TB deaths by 90% by 2030 [@WHO2022c].
+Previous research has focused on trends in TB outcomes,; this project will explore and analyze the disparities within the indicators based on subgroup and burden of disease as well as analyizing trends in tuberculosis outcomes.
 
@@
 ## Questions/Hypotheses to be addressed
+*update depending on ability to bind indicators with outcomes*
 
@@
 In addition to the inequality and indicator predictors provided with the dataset, I would like to examine differences in TB outcomes based on level of TB burden.
-I would need to add another classification based on the literature to indicate what levels of TB incidence and prevalence would fall into each level of disease burden.
+This can be completed by subsetting the high burden countries into a smaller data set.
 
@@
 
-To analyze, I would need to determine how to handle the missing data.
-The years between countries are not specific, and not all indicators were collected for each country.
-At the moment, I would subset the data based on indicator and remove incomplete observations to analyze each indicator individually.
-For the final analysis, I will most likely focus on TB indicators that are significantly different between subgroups or complete enough for further analysis.
-As the class progresses, I look forward to learning other analysis techniques.
-Eventually, I would like to create a dataset that can produce the statistical information, plots, and models that explore TB outcomes based on inequality measures, TB indicators, and/or level of burden.
+The data cleaning will consist of widening the data set so each TB indicator is placed in a column in order to explore the indicators separately.
+The data set will also be split into a full data set containing all countries (wide_data) and high burden countries (highburden).
+The data will then be summarized and described using exploratory analysis approaches such as graphs, plots, and tables.
+The statistical analysis will use cross-validation and decision tree modeling to model TB outcomes based on the available predictor variables.
+A similar approach will be used for the health equity indicators as outcomes.
 
@@
 
-For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator.
+For the statistical analysis of the data, I created decision tree models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator and tuned the models using cross-validation folds.
+The data was initially split into test/train data sets to better evaluate model performance.
 An example is shown below.
@@
 
-Other indicators were also analyzed and fit to a linear model as shown below.
+Other indicators were also analyzed and fit to a decision tree model using the same methods.
 
@@
 
-In the next step of my project, I would like to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes.
-This will most likely be limited to the high burden countries.
+I would still like to try to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes. This will most likely be limited to the high burden countries.
 
@@
 
-*(Part 2 of project: exploratory analysis)* Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
+Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
 
@@
 Most indicators showed the greatest disparities due to economic status and education level.
-Further exploratory analysis should look into level of TB burden as a possible variable that should be considered, and statistical analysis in the next step should attempt to quantify the disparities between subgroups and the impact on TB outcomes.
 
 ## Basic statistical analysis
-
-*(Part 3 of project: linear model fitting and predictions, more to be added)* TBA: simple statistical models for associations in data
+*update depending on binding outcomes and predictors: will probably do linear model predictions from TB outcomes from one indicator if possible. otherwise will do line*
 
@@
 
-TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here
+For the majority of the statistical analysis, the tuberculosis outcomes and the health equity indicators with high disparity between subgroups (as identified during the exploratory analysis) were modeled and tuned using the decision tree model and cross-validation methods from the `tidymodels` guide.
+The models were all used the training data to tune the cost_complexity and tree_depth parameters to find the model with the lowest value of RMSE.
+An example output is shown below.
 
+```{r}
+#tree plot models for full data TB incidence
+readRDS(file = here("results", "tune plots", "wdinctreeplot.rds")) 
+#tree plot models for high burden TB incidence
+readRDS(file = here("results", "tune plots", "hbinctreeplot.rds"))
+```
+
+Please refer to the `statistical_analysis.qmd` for the full code and results.
+
+A summary of the model performances are shown below.
+The full data set models are marked with 'wd' and the high burden models are marked as 'hb'.
+The outcomes being modeled are abbreviated in the `model` column.
+The best performing model predicted prevalence to notification ratio for the full data set (RMSE = 0.971).
+The TB outcome models all performed poorly in predicting TB incidence, mortality, and prevalence.
+This is most likely due to the only significant predictor being country.
+```{r}
+readRDS(file = here("results", "summaryrmse.rds")) #summary table of RMSE from decision tree models
+```
+
+The second research question was to explore the differences between the full data set and high burden countries.
+I hypothesized that the high burden models would have a higher performance because these countries have the highest concentration of TB cases.
+However, the high burden models consistently preformed worse than the full data models.
+This is most likely due to the lower amount of available data.
 {{< pagebreak >}}

### 2023-04-07 code/analysis_code/statistical_analysis.R

@@ -21,6 +21,103 @@ data_location <- here::here("data","processed_data","processeddata.rda")
 #load data. 
 load(data_location)
 
+## ---- basicmodels ---------
+# fit linear model with TB incidence as outcome, subgroup as predictor
+lm_mod <- linear_reg()
+
+#fit model with high burden countries
+inclm_fit <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
+      data = highburden) #fit linear model
+incfittable <- tidy(inclm_fit) 
+print(incfittable) #produce tidy table of fitted model
+
+table_file1 = here("results", "model fit tables", "incfittable.rds")
+saveRDS(incfittable, file = table_file1) #save summary table
+
+#fit model with full data
+inclm_fitFD <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
+      data = wide_data) #fit linear model
+incfittable2 <- tidy(inclm_fitFD) 
+print(incfittable2) #produce tidy table of fitted model
+
+table_file2 = here("results", "model fit tables","incfittableFD.rds")
+saveRDS(incfittable2, file = table_file2) #save summary table
+
+#prediction model using subgroup and setting
+predinclm_fit <- lm_mod %>% 
+  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
+      data = highburden) #fit linear model
+predincfittable <- tidy(predinclm_fit) 
+print(predincfittable)
+
+new_points <- expand.grid(subgroup = c("Female", "Male"),
+                          setting = "India") #predicting Indian incidence based on sex
+incpred <- predict(predinclm_fit,
+                   new_data = new_points) #predictions
+CIincpred <- predict(predinclm_fit,
+                     new_data = new_points,
+                     type = "conf_int") #confidence intervals for predictions
+preddata <- new_points %>% #create plot data
+  bind_cols(incpred) %>% 
+  bind_cols(CIincpred)
+
+predplot <- ggplot(preddata, aes(subgroup))+
+  geom_point(aes(y = .pred))+ #plot predictions with confidence intervals
[…]

### 2023-04-21 products/manuscript/Manuscript.qmd
Part 5 of project
cleaned up manuscript, removed extra text, fixed graphs/plots

@@
 csl: ../apa.csl
+execute: 
+  echo: false
 ---
@@
 ```{r, echo=FALSE, message=FALSE}
+#| output: false
 # load a few R packages
@@
 library(tidyverse)
-library(dplyr)
-library(ggplot2)
-library(tidymodels)
 ```
-
-# Questions for Dr. Handel
-I would like to be able to predict the main TB outcomes based on the health equity indicators (ex: level of TB incidence predicted from % of people with drug-resistant strains), but I cannot join the rows together correctly.
-When I tried bind_rows() on the one of the countries to see if it would work, the outcome rows are still separated and if I remove year to force everything onto the same row, I have 90 variables and they are grouped within the same cell.
-I added an example below.
-Do you have any advice?
-
-```{r}
-#path to data
-data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")
-
-#load data and assign to rawdata
-rawdata <- readxl::read_excel(data_location)
-
-#without year removed: outcomes remain on a separate row
-test <- rawdata %>% 
-  select(c(setting, year, indicator_name,
-           dimension, subgroup,
-           estimate)) %>% 
-  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
-  filter(setting == "Afghanistan") %>% bind_rows()
-test
-
-#with year removed: cells contain multiple observations
-test2 <- rawdata %>% 
-  select(c(setting, indicator_name,
-           dimension, subgroup,
-           estimate)) %>% 
-  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
-  filter(setting == "Afghanistan") %>% bind_rows()
-test2
-```
-If it is not feasible, I will be analyzing TB outcomes separately from indicators instead.
 
@@
 
-Full summary to be added (TBA): background of topic, objective of project, quick overview of methods, main findings
+Tuberculosis is the leading infectious cause of death, and the burden of TB is disproportionately concentrated in several high burden countries.
+The reason for the disparity between countries and populations of specific demographics is not fully understood.
+The World Health Organization collects annual data on tuberculosis outcomes through the Global Tuberculosis Programme with the goal of reducing TB incidence worldwide.
+The further explore the reasons behind the health inequity surrounding tuberculosis, this analysis explores the health equity indicators collected by WHO and attempts to predict tuberculosis outcomes.
+
+The project cleans and summarizes the health equity indicators to explore disparities within the indicators based on level of income, education, sex, place of residence, and presence of drug-resistant strains.
+The data set was also subset into high burden countries to explore difference between high burden countries and all countries.
+Linear regression and decision tree modelling using cross-validation tuning were conducted with predict tuberculosis outcomes and level of equity indicators.
+
+In 2020, Lesotho had the highest TB incidence, but with the inclusion of mortality rates, Central African Republic had the highest case fatality rate with a high incidence and mortality.
+Males have a higher proportion of cases compared to females, and level of income and education had the largest disparity between the highest and lowest levels.
+More research needs to be conducted to measure the direct relationship between the health equity indicators and tuberculosis outcomes.
 
@@
 Their current objective is to reduce TB incidence by 80% and TB deaths by 90% by 2030 [@WHO2022c].
-Previous research has focused on trends in TB outcomes,; this project will explore and analyze the disparities within the indicators based on subgroup and burden of disease as well as analyizing trends in tuberculosis outcomes.
+Previous research has focused on trends in TB outcomes; this project will explore and analyze the disparities within the indicators based on subgroup and burden of disease as well as analyzing trends in tuberculosis outcomes.
 
@@
 ## Questions/Hypotheses to be addressed
-*update depending on ability to bind indicators with outcomes*
+*update depending on ability to bind indicators with outcomes; discuss with Zane next week*
 
@@
 This data was downloaded and stored into the `raw_data` folder of the repository as `rawdata`.
-Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`
-
-```{r load data}
-#path to data
-data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")
-
-#load data and assign to rawdata
-rawdata <- readxl::read_excel(data_location)
-
-str(rawdata)
-```
-
-Please see `processingfile.qmd` for full code describing data importing and cleaning.
-
-The `rawdata` file contains information on the country, year of data collection, the indicator they are studying, the social determinants of health they are considering, and the subgroup of each dimension.
-The dataset is mostly complete with only a few main variables missing data points.
-The indicators being studied are shown below.
-
-```{r}
-#look at what indicators are in the dataset
-unique(rawdata$indicator_name)
-```
+Information on the indicators, social determinants dimensions, and subgroups can be found in the `raw_data` folder under `202206-metadata-tb.pdf`.
+Please render the `processingfile.qmd` file for full results from data importing and cleaning.
+The associated R script file contains all coding steps with descriptions.
 
@@
 These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup.
-An example of the data cleaning is shown below.
-
-```{r}
-wide_data <- rawdata %>% 
-  select(c(setting, year, indicator_name,
-           indicator_abbr, dimension, subgroup,
-           estimate, population)) %>% 
-  pivot_wider(names_from = "indicator_name", values_from = "estimate")
-head(wide_data)
-
-# BCG coverage indicator
-wide_data %>% 
-  filter(indicator_abbr == "bcg") %>% 
-  select(c(1,2,3,4,5,6,7)) %>% 
-  summary()
-```
 
@@
 
-```{r}
-#path to data
-data_location <- here::here("data","processed_data","processeddata.rda")
-#load data. 
-load(data_location)
-```
-
 ## Statistical analysis
@@
 The data was initially split into test/train data sets to better evaluate model performance.
-An example is shown below.
 This method was repeated for the `wide_data` and `highburden` countries.
 
-```{r}
-# fit linear model using TB incidence as outcome, subgroup as predictor
-lm_mod <- linear_reg() #set linear regression
-
-#fit model with high burden countries
-inclm_fit <- lm_mod %>% 
-  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
-      data = highburden) #fit linear model
-incfittable <- tidy(inclm_fit) 
-print(incfittable) #produce tidy table of fitted model
-```
-
-Other indicators were also analyzed and fit to a decision tree model using the same methods.
-
-```{r}
-# fit linear model with BCG coverage as outcome, subgroups as predictors
-
-#fit model with high burden countries
-BCGlm_fit <- lm_mod %>% 
-  fit(`BCG immunization coverage among one-year-olds (%)` ~ subgroup,
-      data = highburden) #fit linear model
-BCGfittable <- tidy(BCGlm_fit) 
-print(BCGfittable) #produce tidy table of fitted model
-```
-
-I would still like to try to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes. This will most likely be limited to the high burden countries.
+I would still like to try to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes.
+This will most likely be limited to the high burden countries.
+I am meeting with Zane next week to discuss.
 
@@
 
-Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.
-
-The `processedcode` was loaded into the `exploratory_analysis.qmd` file which pulls coding from the `exploratoryanalysis.r` script.
+After processing the data, the `wide_data` and `highburden` datasets were described by exploring each health equity indicators and tuberculosis outcome.
+@fig-indicators shows the list of indicators, and the codebook can be found in the README.md file.
 
 ```{r}
-#Path to data.
-data_location <- here::here("data","processed_data","processeddata.rda")
-#load data
-load(data_location)
+#| label: fig-indicators
+#| fig-cap: "Healthy equity indicators"
+readRDS(file = here("results", "processing", "indicators.rds"))
 ```
@@
 The summary tables are stored in the `results` folder.
-An example of the code is shown below.
+The population for most variables were relatively low and skewed to the right.
+The TB incidence and mortality indicators also showed a skewed distribution.
+The highest incidence rate was 908 cases/100,000 shown in @fig-incsum, and highest mortality rate was 140 deaths/100,000 shown in @fig-mortsum.
+```{r}
+#| label: fig-incsum
+#| fig-cap: "Tuberculosis incidence summary"
+readRDS(file = here("results", "exploratory_analysis", "summary tables", "incsummarytable.rds"))
+```
 
 ```{r}
-#incidence summary and save to file location
-summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
-print(summary_inc)
-
-#mortality summary and save to file location
-summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
-print(summary_mort)
+#| label: fig-mortsum
+#| fig-cap: "Tuberculosis mortality summary"
+readRDS(file = here("results", "exploratory_analysis", "summary tables", "mortsummarytable.rds"))
 ```
 
-The population for most variables were relatively low and skewed to the right.
-The TB incidence and mortality indicators also showed a skewed distribution.
-The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.
-
-After getting an overview of the indicators, I wanted to explore the differences between subgroups and start identifying the largest disparities.
+After getting an overview of the indicators, I explored the differences between subgroups within each health equity indicator and began identifying the largest disparities.
 Using `ggplot()`, I created plots of the indicators separated by dimension and subgroup.
-For most variables, these disparities were described using the `boxplot()` function as shown below.
+For most variables, these disparities were described using the `boxplot()` function.
+The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function.
 
 ```{r}
-p13 <- highburden %>%
-  filter(!dimension %in% c("Age (2 groups) (0-15+)", "TB drug resistance", "Sex")) %>%
-  ggplot(aes(x=dimension,
-             y = `People who report TB is spread through coughing - Female (%)`,
-             fill = subgroup)) +
-  geom_boxplot()+
-  theme(axis.text.x = element_text(angle = 90))   
-plot(p13) # plots knowledge about TB by dimension and subgroup for females
+#| label: fig-incexplore
+#| fig-cap: "Tuberculosis incidence by country"
+include_graphics(here("results", "exploratory_analysis", "figures", "incexplore1.png"))
+```
+@fig-incexplore shows the tuberculosis incidence for each country stratified by sex.
+Males have a higher proportion of TB cases, and most countries have minimal cases per 100,000 people.
+@fig-incexplorehb shows the TB incidence for high burden countries which are responsible for most of the TB burden worldwide.
+Lesotho has the highest number of cases followed by South Africa and Central African Republic.
+
+```{r}
+#| label: fig-incexplorehb
+#| fig-cap: "Tuberculosis incidence by country for high burden countries"
+include_graphics(here("results", "exploratory_analysis", "figures", "incexplore2.png"))
 ```
 
-The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
+@fig-mortexplore shows the TB mortality for high burden countries. Compared with the TB incidence values, Central African Republic has high TB incidence and mortality. 
+```{r}
+#| label: fig-mortexplore
+#| fig-cap: "Tuberculosis mortality by country"
+include_graphics(here("results", "exploratory_analysis", "figures", "mortexplore.png"))
+```
 
+@fig-prevexplore shows tuberculosis prevalence by country. There may be a slight trend of urban areas having higher prevalence compared to rural areas.
 ```{r}
-p6 <- highburden %>% 
-  filter(dimension %in% "Sex") %>%
-  filter(indicator_abbr == "mortality") %>% 
-  select(c(1,2,3,4,5,6,13)) %>% 
-  ggplot(aes(x=fct_reorder(
-  setting, `TB mortality (deaths per 100 000 population)`),
-  y = `TB mortality (deaths per 100 000 population)`,
-  color = subgroup)) +
-    geom_point()+
-  theme(axis.text.x = element_text(angle = 90))+
-  scale_fill_brewer(palette = "Spectral")+
-  xlab("setting")
-plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
+#| label: fig-prevexplore
+#| fig-cap: "Tuberculosis prevalence by country"
+include_graphics(here("results", "exploratory_analysis", "figures", "prevexplore1.png"))
+```
+
+When exploring the health equity indicators, the BCG immunization proportion, catastrophic costs due to TB, prevalence to notification ratio, and knowledge about TB showed the largest disparities between subgroups.
+@fig-bcg shows the percentage of immunization for one-year-olds based on economic status, education level, area of residence, and sex.
+The economic status and education levels showed an approximate difference of 10-15% vaccination coverage between the highest and lowest levels.
+```{r}
+#| label: fig-bcg
+#| fig-cap: "BCG immunization by subgroup"
+include_graphics(here("results", "exploratory_analysis", "figures", "bcgexplore.png"))
+```
+Experiencing catastrophic costs due to TB was most heavily present in the lowest economic quintile and was highly associated with drug resistant TB with approximately 85% of people with drug-resistant strains experiencing high costs.
+```{r}
+#| label: fig-catacost
+#| fig-cap: "Catastropic costs due to TB in highburden countries"
+include_graphics(here("results", "exploratory_analysis", "figures", "catacostexplore.png"))
+```
+
+The prevalence to notification also shows a mild inequity between females and males as shown in @fig-prevnot.
+Males on average have 2.4 years between contracting tuberculosis and receiving a diagnosis compared to 1.7 years for females.
+```{r}
+#| label: fig-prevnot
+#| fig-cap: "Time between prevalence to notification in high burden countries"
+include_graphics(here("results", "exploratory_analysis", "figures", "pnexplore.png"))
+```
+
+The final indicator that showed the highest disparity is the male and female knowledge of TB shown in @fig-knowledge.
+Both groups show similar disparities within the economic status, education level, and place of residence subgroups.
+However, females appear to have more extreme disparities within the subgroups.
+For example, males in the lowest quintile have 60% of the population exhibiting knowledge about how TB is spread, but females of the same quintile show only 50% of population having TB knowledge.
+Similar trends occur for th education and place of residence subgroups.
+```{r}
+#| label: fig-knowledge
+#| fig-cap: "Knowledge about TB by subgroup for each sex"
+#| fig-show: hold
+#| fig-align: center
+include_graphics(c(here("results", "exploratory_analysis", "figures", "coughmexplore.png"), here("results", "exploratory_analysis", "figures", "coughfexplore.png")))
 ```
@@
 Most indicators showed the greatest disparities due to economic status and education level.
+Please render the `exploratory_analysis.qmd` file for the full exploratory analysis exploring the other health equity indicators.
 
 ## Basic statistical analysis
-*update depending on binding outcomes and predictors: will probably do linear model predictions from TB outcomes from one indicator if possible. otherwise will do line*
+*update depending on binding outcomes and predictors: will probably do linear model predictions from TB outcomes from one indicator if possible.*
 
-The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
+The linear models summary for TB incidence are listed below as @tbl-incfithb and @tbl-incfit.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
 The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`.
-Based on the p-value for the subgroup variable, sex has a significant association with TB incidence in both models.
 Males having higher TB incidence is congruent with the exploratory analysis findings.
@@
 ```{r}
-readRDS(file = here("results", "model fit tables", "incfittable.rds")) #TB incidence model for high burden countries
-readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) #TB incidence model for full data
+#| label: tbl-incfithb
+#| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
+readRDS(file = here("results", "model fit tables", "incfittable.rds")) 
+```
+```{r}
+#| label: tbl-incfit
+#| fig-cap: "Summary table for fitted linear model predicting incidence"
+readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) 
 ```
 
-The fitted model summaries below are the TB mortality for high burden countries and full data.
+The fitted model summaries below are the TB mortality for high burden countries (@tbl-mortfithb) and full data (@tbl-mortfit).
 For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
-Sex is a statistically significant predictor of TB mortality (p-values \<0.05), but this does not consider any other factors.
-The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators.
 
 ```{r}
-readRDS(file = here("results", "model fit tables", "mortfittable.rds")) #TB mortality model for high burden countries
-readRDS(file = here("results", "model fit tables", "mortfittableFD.rds")) #TB mortality model for full data
+#| label: tbl-mortfithb
+#| fig-cap: "Summary table for fitted linear model predicting mortality for high burden countries"
+readRDS(file = here("results", "model fit tables", "mortfittable.rds"))
+```
+```{r}
+#| label: tbl-mortfit
+#| fig-cap: "Summary table for fitted linear model predicting mortality"
+readRDS(file = here("results", "model fit tables", "mortfittableFD.rds"))
 ```
 
-The last outcome model I fitted was TB prevalence for high burden countries and the full data.
+The last outcome model I fitted was TB prevalence for high burden countries (@tbl-prevfithb) and the full data (@tbl-prevfit).
 For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0.
@@
 ```{r}
-readRDS(file = here("results", "model fit tables", "prevfittable.rds")) #TB prevalence model for high burden countries
-readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) #TB prevalence model for full data
+#| label: tbl-prevfithb
+#| fig-cap: "Summary table for fitted linear model predicting prevalence for high burden countries"
+readRDS(file = here("results", "model fit tables", "prevfittable.rds"))
+```
+```{r}
+#| label: tbl-prevfit
+#| fig-cap: "Summary table for fitted linear model predicting prevalence"
+readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) 
 ```
 
-The models can be further explored by including `setting` as a predictor, shown in the prediction model below.
+The models can be further explored by including `setting` as a predictor, shown in the prediction model below as @fig-prediction.
 The prediction model is basic prediction of TB incidence from subgroup and setting.
-As an example, I predicted the incidence for females and males in India (shown in the plot).
-The estimates for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) are shown using `geom_point()` and `geom_errorbar()`.
-I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables.
+As an example, I predicted the incidence for females and males in Central African Republic (shown in the plot).
+The estimates for males (602.5720; CI = 492.8893, 712.2546) and females (476.3097; CI = 366.6271, 585.9924) are shown using `geom_point()` and `geom_errorbar()`.
 
 ```{r}
-#prediction model for TB incidence with subgroup and setting as predictors
-predinclm_fit <- lm_mod %>% 
-  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
-      data = highburden) #fit linear model
-predincfittable <- tidy(predinclm_fit) 
-print(predincfittable)
-
-#prediction plot using the above model estimating TB incidence for females and males in India
+#| label: fig-prediction
+#| fig-cap: "Prediction plot estimating TB incidence for females and males in Central African Republic"
 readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
@@
 The models were all used the training data to tune the cost_complexity and tree_depth parameters to find the model with the lowest value of RMSE.
-An example output is shown below.
+
+The @tbl-treeinc displays an example of the model tuning for decision tree model predicting incidence, and @tbl-treeinchb shows the same method for high burden countries.
+Please refer to the `statistical_analysis.qmd` for the full code and results on modelling.
+```{r}
+#| label: tbl-treeinc
+#| fig-cap: "Summary table for fitted tree model predicting incidence"
+readRDS(file = here("results", "tune plots", "wdinctreeplot.rds")) 
+```
 
 ```{r}
-#tree plot models for full data TB incidence
-readRDS(file = here("results", "tune plots", "wdinctreeplot.rds")) 
-#tree plot models for high burden TB incidence
+#| label: tbl-treeinchb
+#| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
 readRDS(file = here("results", "tune plots", "hbinctreeplot.rds"))
@@
 
-Please refer to the `statistical_analysis.qmd` for the full code and results.
-
-A summary of the model performances are shown below.
+A summary of the model performances are shown below predicting TB outcomes and healthy equity indicators by subgroup and setting.
 The full data set models are marked with 'wd' and the high burden models are marked as 'hb'.
@@
 ```{r}
-readRDS(file = here("results", "summaryrmse.rds")) #summary table of RMSE from decision tree models
+#| label: tbl-rmsesum
+#| fig-cap: "Summary table of RMSE values for tree models"
+readRDS(file = here("results", "summaryrmse.rds"))
 ```
@@
 
+The objective of this project was to compare disparities in the health equity indicators with tuberculosis outcomes to determine if larger disparities are associated with tuberculosis burden.
+Due to the formatting of the data, the relationship could not be directly compared.
+Instead, each equity indicator was individually analyzed, and the tuberculosis outcomes were modeled using the countries and subgroups.
+
 ## Summary and Interpretation
 
-TBA: summary of findings and why it matters
+The data for this analysis was collected by the World Health Organization to study tuberculosis outcomes and explore disparities in several underlying indicators of health equity including knowledge about how tuberculosis is spread, proportion of children vaccinated against TB, and percentage of people that experience catastrophic costs due to the disease.
+The health equity indicators can be viewed as potential targets of public health programs in order to reduce the morbidity of TB.
+
+Most countries in the world experience relatively low levels of tuberculosis cases, so very few countries experience high burden of tuberculosis.
+In 2020, Lesotho had the highest number of TB cases (908 cases/100,000 population) followed by South Africa and Central African Republic.
+When considering mortality of the same year, Central African Republic had the highest mortality, indicating the country had the highest case fatality rate.
+
+Males tended to have a higher proportion of the tuberculosis cases when exploring the incidence, prevalence, and mortality.
+In addition, males had a longer prevalence to notification ratio compared to females.
+There are many factors that may be contributing to the greater TB burden on males, this delay in notification may be associated with longer transmission time which could be a factor in the higher incidence.
+
+The initial statistical analysis predicted tuberculosis outcomes using a linear model.
+The incidence and mortality models were consistent with the exploratory analysis since being male was the main predictor of the model.
+The high burden countries have higher levels of TB outcomes compared to the full data set.
+However, these models did not take into account the country as a predictor, so the full analysis used a decision tree model to include this as a predictor.
+
+The full analysis consisted of a series of decision tree models that predicted TB outcomes and equity indicators by subgroups.
+The high burden models were predicted to perform better than the full set models, but this hypothesis did not hold true.
+All tuberculosis outcome models did not perform well with the incidence, mortality, and prevalence models for the high burden countries having RMSE values of 163, 21.6, and 184 respectively.
+The prevalence model was the only outcome in which the high burden model performed better than the full model.
+
+Of the healthy equity indicators, the prevalence to notification ratio, attitudes toward TB, and BCG vaccination coverage models were the top performing models.
+
+For most decision tree models, country was the most significant predictor.
+This indicates that public health programs should personalize their efforts to eliminate tuberculosis using country/population specific factors.
+
+After discussing with Zane, the goal would be to be able to use health equity indicator level (ex: BCG vaccine coverage) to predict TB outcome.
+With this information, you would be able to model how changes would improve TB outcome (if you raise the BCG vaccine coverage to at least 80%, the predicted TB incidence would lower by ##).
 
@@
 
-TBA: strengths and limitations of analysis
+This project had several limitations associated with the analysis.
+Due to the collection years differing between indicator variables and TB outcomes, I am not able to predict tuberculosis incidence, mortality, and prevalence from the health inequity indicators.
+In order to complete this analysis as my research question designed, I would be losing valuable data on the subgroups by having to summarize the overall indicators by country.
+I will be discussing this with Zane and either adjust my research question or reformat my data.
+Another limitation of the analysis is the difficulty interpreting the decision tree models.
+Since country was the most significant predictor for the models, it was difficult to visualize and express the model.
+The final limitation is that the data did not capture all factors that contribute to the spread of tuberculosis such as population density, access to treatment regimens, consistency of adherence, and other aspects that can affect one's susceptibility to contracting or spreading the disease.
+The strengths of the analysis is the substantial data set which accurately described countries of various tuberculosis burden rather than focusing specifically on countries with TB cases.
+The analysis also thoroughly explored each equity indicator to highlight trends and populations that could aid in the distribution of public health resources.
 
 ## Conclusions
-
-TBA: take away messages, citations from bibtex
+This analysis explores tuberculosis-related disparities between levels of income, education, sex, and other health equity topics.
+TB incidence and mortality is most closely associated with the country.
+There is no direct comparison between level of disparity in equity indicator and tuberculosis incidence, prevalence, or mortality.
+Continued analysis is required to address the gap in data preventing a direct measurement of health disparity on tuberculosis outcomes.
+Further research should explore impact of programs addressing the health equity covered during this analysis on tuberculosis incidence and mortality.
 

### 2023-04-21 code/analysis_code/exploratoryanalysis.R

@@ -12,6 +12,7 @@ library(dplyr)
 library(skimr)
 library(ggplot2) #for plots/figures
 library(forcats) #for factoring
+library(scales)
 
 
 ## ---- loaddata --------
@@ -24,145 +25,161 @@ load(data_location)
 #bcg coverage summary and save to file location
 summary_bcg <- skimr::skim(wide_data$`BCG immunization coverage among one-year-olds (%)`)
 print(summary_bcg)
-bcgsummarytable_file <- here("results", "summary tables", "bcgsummarytable.rds")
+bcgsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "bcgsummarytable.rds")
 saveRDS(summary_bcg, file = bcgsummarytable_file)
 
 #catastrophic cost summary and save to file location
 summary_catacost <- skimr::skim(wide_data$`Families affected by TB facing catastrophic costs due to TB (%)`)
 print(summary_catacost)
-catacostsummarytable_file <- here("results", "summary tables", "catacostsummarytable.rds")
+catacostsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "catacostsummarytable.rds")
 saveRDS(summary_catacost, file = catacostsummarytable_file)
 
 #case detection rate summary and save to file location
 summary_cdr <- skimr::skim(wide_data$`Case detection rate (%)`)
 print(summary_cdr)
-cdrsummarytable_file <- here("results", "summary tables", "cdrsummarytable.rds")
+cdrsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "cdrsummarytable.rds")
 saveRDS(summary_cdr, file = cdrsummarytable_file)
 
 #drug resistance summary and save to file location
 summary_drug <- skimr::skim(wide_data$`People with MDR/RR-TB (%)`)
 print(summary_drug)
-drugsummarytable_file <- here("results", "summary tables", "drugsummarytable.rds")
+drugsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "drugsummarytable.rds")
 saveRDS(summary_drug, file = drugsummarytable_file)
 
 ## ---- exploratorytables2 --------
 #incidence summary and save to file location
 summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
 print(summary_inc)
-incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
[…]

### 2023-04-21 code/analysis_code/statistical_analysis.R

@@ -32,7 +32,7 @@ inclm_fit <- lm_mod %>%
 incfittable <- tidy(inclm_fit) 
 print(incfittable) #produce tidy table of fitted model
 
-table_file1 = here("results", "model fit tables", "incfittable.rds")
+table_file1 = here("results", "statistical_analysis", "model fit tables", "incfittable.rds")
 saveRDS(incfittable, file = table_file1) #save summary table
 
 #fit model with full data
@@ -42,7 +42,7 @@ inclm_fitFD <- lm_mod %>%
 incfittable2 <- tidy(inclm_fitFD) 
 print(incfittable2) #produce tidy table of fitted model
 
-table_file2 = here("results", "model fit tables","incfittableFD.rds")
+table_file2 = here("results", "statistical_analysis", "model fit tables","incfittableFD.rds")
 saveRDS(incfittable2, file = table_file2) #save summary table
 
 #prediction model using subgroup and setting
@@ -53,7 +53,7 @@ predincfittable <- tidy(predinclm_fit)
 print(predincfittable)
 
 new_points <- expand.grid(subgroup = c("Female", "Male"),
-                          setting = "India") #predicting Indian incidence based on sex
+                          setting = "Central African Republic") #predicting CAR incidence based on sex
 incpred <- predict(predinclm_fit,
                    new_data = new_points) #predictions
 CIincpred <- predict(predinclm_fit,
@@ -68,10 +68,11 @@ predplot <- ggplot(preddata, aes(subgroup))+
   geom_errorbar(aes(ymin = .pred_lower,
                     ymax = .pred_upper),
                 width = .2) +
-  labs(y = "TB Incidence (per 100 000 population)")
+  labs(y = "TB Incidence (per 100 000 population)", 
+       title = "Predicted TB incidence for Central African Republic")
 predplot
 
-saveRDS(predplot, file = here("results", "model fit tables", "prediction1.rds"))
+saveRDS(predplot, file = here("results", "statistical_analysis", "model fit tables", "prediction1.rds"))
 
 # fit linear model using TB mortality as outcome, subgroup and setting as predictor
 
[…]

### 2023-04-21 code/processing_code/processingcode.R

@@ -49,7 +49,9 @@ length(unique(rawdata$setting))
            indicator_abbr, dimension, subgroup,
            estimate, population)) %>% 
   pivot_wider(names_from = "indicator_name", values_from = "estimate")
-slice(wide_data)
+summarywide <- tbl_sum(wide_data)
+data1 = here("results", "processing", "summarywide.rds")
+saveRDS(summarywide, file = data1)
 
 #subset high burden countries
 highburden <- wide_data %>% 
@@ -71,10 +73,15 @@ highburden <- wide_data %>%
            "Somalia", "Tajikistan", "Ukraine",
            "Uzbekistan", "Zimbabwe")) 
 unique(highburden$setting)
+summaryhb <- tbl_sum(highburden)
+data2 = here("results", "processing", "summaryhighburden")
+saveRDS(summaryhb, file = data2)
 
 ## ---- cleandata2 --------
 # look at each indicator to check for cleaning requirements
-unique(wide_data$indicator_abbr)
+indicator <- unique(wide_data$indicator_abbr)
+fig2 = here("results", "processing", "indicators.rds")
+saveRDS(indicator, file = fig2)
 
 # BCG coverage indicator
 wide_data %>% 

### 2023-04-28 products/manuscript/Manuscript.qmd
Project Review-Jacob Matta

@@
 #| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
-readRDS(file = here("results", "model fit tables", "incfittable.rds")) 
+readRDS(file = here("results", "statistical_analysis", "model fit tables", "incfittable.rds")) 
 ```
+
 ```{r}
@@
 #| fig-cap: "Summary table for fitted linear model predicting incidence"
-readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) 
+readRDS(file = here("results", "statistical_analysis","model fit tables", "incfittableFD.rds")) 
 ```
@@
 #| fig-cap: "Summary table for fitted linear model predicting mortality for high burden countries"
-readRDS(file = here("results", "model fit tables", "mortfittable.rds"))
+readRDS(file = here("results", "statistical_analysis", "model fit tables", "mortfittable.rds"))
 ```
@@
 #| fig-cap: "Summary table for fitted linear model predicting mortality"
-readRDS(file = here("results", "model fit tables", "mortfittableFD.rds"))
+readRDS(file = here("results","statistical_analysis", "model fit tables", "mortfittableFD.rds"))
 ```
@@
 #| fig-cap: "Summary table for fitted linear model predicting prevalence for high burden countries"
-readRDS(file = here("results", "model fit tables", "prevfittable.rds"))
+readRDS(file = here("results","statistical_analysis", "model fit tables", "prevfittable.rds"))
 ```
@@
 #| fig-cap: "Summary table for fitted linear model predicting prevalence"
-readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) 
+readRDS(file = here("results", "statistical_analysis", "model fit tables", "prevfittableFD.rds")) 
 ```
@@
 #| fig-cap: "Prediction plot estimating TB incidence for females and males in Central African Republic"
-readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
+readRDS(file = here("results", "statistical_analysis", "model fit tables", "prediction1.rds")) 
 ```
@@
 #| fig-cap: "Summary table for fitted tree model predicting incidence"
-readRDS(file = here("results", "tune plots", "wdinctreeplot.rds")) 
+readRDS(file = here("results", "statistical_analysis","tune plots", "wdinctreeplot.rds")) 
 ```
@@
 #| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
-readRDS(file = here("results", "tune plots", "hbinctreeplot.rds"))
+readRDS(file = here("results", "statistical_analysis","tune plots", "hbinctreeplot.rds"))
 ```

### 2023-04-28 code/processing_code/processingcode.R

@@ -15,6 +15,7 @@ library(tidyr) #for data processing/cleaning
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
 library(naniar) #to look for missing data
+library(pillar)
 
 ## ---- loaddata --------
 #path to data

### 2023-04-28 review_Jacob_Matta.qmd

---
  title: Project Review Template 
author: 
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
  toc: true
toc-depth: 3
number-sections: true
---

# Overview

Title of project: sarabenist-MADA-project

Name of project author(s): Sara Benist

Name of project reviewer: Jacob Matta

## HELLO

Hi Sara, I am happy to be reviewing your project as it appears to be quite organzied and complete. Please bear with me as I attempt to provide meaningful comments in order to help you further develop your project. I thank you for your time in advance!

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Review of Summary/Abstract
The first two sentences of your Summary/Abstract section are quite important and strong. Immediately the reader is aware "the burden of TB" in "high burden countries" is the topic of your research. You then mention that there is a lack of information understanding why there are disparities between countries and populations when it comes to the burden of TB. From these two sentences I immediately understand what you will be conducting your anaylsis on and why there is a need for the analysis, which is great information to know as a reader.

In your fourth sentence of the first paragraph in the Summary/Abstract "**The** further explore the reasons behind the health inequity surrounding tuberculosis, this analysis explores the health equity indicators collected by WHO and attempts to predict tuberculosis outcomes" I think you meant to use the word "To" instead of "The" to start the sentence. Other than that, I like this sentence because you directly tell the reader what the focus of your analysis will be.

In second paragraph, gives a nice introduction into the analysis you conducted which appears to include: cleaning and exploring by health equity indicators, subsetting the data by high burden countries, and then doing linear regression followed by decision tree modeling to predict TB outcomes. This is a strong paragraph that helps the reader understand how you will be conducting the analysis.

The third paragraph provides some further context and known information of TB disparities between countries. It may be useful to add a little more information on the two countries you use in this paragraph, both Lesotho and Central African Republic. Adding information regarding where these countries are in the world would be helpful for me at least(I'm someone with awful geographical knowledge). I assume Central African Republic is in Central Africa but I have no idea where Lesotho is. Other than that, this paragraph does a great job giving the reader an idea to some of the health equity indicators like gender, income and education.

### Review of Introduction
The introduction provides further background regarding TB which is helpful for me because I do not know much about TB. Your second sentence is quite powerful "Approximately 10 million people are infected each year, and 1.5 million die from the disease" as it is staggering to me that 1.5 million people die from TB when about 10.5 million people are infected. You also mention what the WHO is doing to investigate the disparities and their goals by 2030. You provide the relevant context for the reader to understand the severity of TB throughout the world and this puts emphasis on why your analysis is important.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
-   strong contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

It seems like you have a 6 questions you are investigating through this analysis:

1. "Does high disparities between inequality measures show high correlation to TB incidence, prevalence, and mortality of a country?"

2. "How do these relationships differ for high burden countries compared to all countries?"

3. "I would like to be able to identify populations that could be a focus of TB health improvement programs"

4. "I would like to explore include regional differences in drug-resistant TB"

5. "the attitudes and perceptions for high burden areas compared to medium or low burden regions"

6. "I would like to examine differences in TB outcomes based on level of TB burden"

All these appear to relate to the data and are able to be addressed. I have yet to explore the results and discussion section as I write this, but I am interested to see the results for these questions. Regarding question 5, there is little background regarding the attitudes and perceptions of TB in countries in your introduction. I do think question 5 is relevant and important, but a little background on the perceptions and attitudes of TB in countries may be helpful for the reader to understand why it is of interest.

### Summary assessment
-   question/hypotheses fully clear

## Data description

You did a good job describing the data as I understand that it was collected from "WHO Global TB programme, TB prevalence surveys, country-specific TB programmes, the WHO Health Equity Monitor database, TB patient cost surveys, and other sources".

When using the link https://www.who.int/data/inequality-monitor/data#PageContent_C158_Col00 to find the data set, I actually had a difficult time finding the "Tuberculosis Indicators" section. However, I was able to find the data set under the "Tuberculosis (multiple data sources)" section which contained a data set with 10 indicators and 194 countries which aligns with your description of the data set. You include a code book for the data which is helpful as well.

### Summary assessment
-   source and overall structure of data well explained

## Data wrangling and exploratory analysis

Your cleaning/processing code is well explained in the manuscript. I was unable to render the `processingfile.qmd` as the error that results is

"Quitting from lines 59-59 (processingfile.qmd) 
Error in tbl_sum(wide_data) : could not find function "tbl_sum"
Calls: .main ... withVisible -> eval_with_user_handlers -> eval -> eval

Execution halted"

It seems that the function pillar::tbl_sum was not found because I do not believe you included the pillar package in your library. Once I included  library(pillar), the file was able to run. Your code seems to follow the steps you list in the manuscript. You don't describe your cleaning/processing in the supplementary file but I'm not sure you need to as you describe the main data cleaning steps in the manuscript.

### Summary assessment
-   essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Exploratory Analysis 
You summarized each each health equity indicator and tuberculosis outcome and you mention each table is in the results folder. You may want to mention that all the tables are in the `summary tables` folder and also provide a `readme` file that briefly describes each table. This would be helpful for those interested in teh summary tables. The summary tables are self explanatory by name but an additional `readme` file is always helpful when navigating a folder.

After you summarized the indicators in each group you briefly discussed the results which is helpful and helps viewer understand the descriptive analysis better.

Next, ggplot what used to create plots of the indicators in order to get a sense of the differences between subgroups. Specifically you gave insight to TB incidence by country, tb incidence by country for high burden countries, tb mortality by country, tb prevalence by country, BCG immunization by subgroup, Catastrophic costs due to TB in high burden countries, time between prevalence to notification in high burden countries, and knowledge about TB by subgroup for each sex while further analysis was provided in the `exploratory_analysis.qmd` file.

All of descriptive analysis appears to be done properly and the results were summarized well in the Manuscript, as the `exploratory_analysis.qmd` file rendered without any issues. I also went through the code and it was well documented as I could follow along.

### Statistical Analysis

#### bivariate analysis:

1. You first ran linear regression for high burden countries and all countries with an outcome of Tb incidence and predictor of subgroupMale and the results were discussed.

2. Linear regression for high burden countries and all countries with an outcome of Tb mortality and predictor of subgroupMale and the results were discussed.

3. Linear regression for high burden countries and all countries with an outcome of TB Prevalence  and predictor of subgroupUrban and the results were discussed.

I think it would be helpful to state why you chose these outcomes and predictors and performed linear regression here. The results are fascinating and do align with your research questions. My concern is why did you chose to share these three bivariate analyses when there a bunch of predictors to choose from.

#### Full Analysis 
I do not have much to comment on regarding your full analysis as it was well explains and documented. Great job.

### Summary assessment
-   strong and reasonable analysis

## Presentation

The overall presentation of your manuscript is very organized and clean in respect to the amount of information you share with the reader. I like your tables in the exploratory/descriptive analysis section of your Manuscript as they are easy to follow for me. However, one of the questions I am asked to consider as a reviewer is "Are the main figures/tables publication level quality?". The tables in the exploratory/descriptive section appear to be R output, as this is what I would have done if I was in your situation. If you have the time it may be worth it to see if there is a way to organize a clean table through R that isn't just output. This is something I will look into as well as I am unsure if this is even possible. If I do find that there is a way to organize tables in R, I will make sure to reach out.

In Figure 1, the countries on teh x-axis are not visible. I can see that this is a tough issue to resolve because there are so many countries to include. Possibly shrinking the text may help?

Great job on the presentation!

### Summary assessment
-   results are very well presented

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

What was discussed:
1. most countries experience low levels of TB

2. Lesotho had the highest number of TB cases (908 cases/100,000 population) followed by South Africa and Central African Republic

3. Central African Republic had the highest mortality rate due TB

4. Males tended to have a higher proportion of the tuberculosis cases when exploring the incidence, prevalence, and mortality

5. Males had a longer prevalence to notification ratio compared to females.

6. The high burden countries have higher levels of TB outcomes compared to the full data set.

7. The full analysis consisted of a series of decision tree models that predicted TB outcomes and equity indicators by subgroups

8. results of the full analysis were discussed

9. Machine learning models were discussed

10. Strength and limitations were discussed

In the discussion and conclusion section you do a good job summarizing the methods you used and the results. In regards to your research questions, I think you may have to go back and make sure some of those are answered more fully. Questions like  "I would like to be able to identify populations that could be a focus of TB health improvement programs",  "I would like to explore include regional differences in drug-resistant TB" and "the attitudes and perceptions for high burden areas compared to medium or low burden regions" may have been answered through exploratroy and statistical anaylsis but they were not discussed in the discussion and conclusion.

### Summary assessment
-   minor parts wrong, missing or unclear

# Overall project content evaluation

Overall your project is very interesting, organized and well documented. You may want to specify your research question as the questions/hypotheses you intend to address are not covered in the discussion and conclusion. I do think you answer most of the questions through your analysis so by discussing the results that directly answer "I would like to be able to identify populations that could be a focus of TB health improvement programs",  "I would like to explore include regional differences in drug-resistant TB" and "the attitudes and perceptions for high burden areas compared to medium or low burden regions" in your discussion would make your discussion stronger. If you do not think you can answer all the questions you originally set out to answer, I think that is okay but make sure to mention that in the discussion as well.

## Structure

Very well organized! Easy to follow!

### Summary assessment
-   well structured

## Documentation

Your project is well documented as you help guide the reader along through the code for you cleaning and analysis. It could be helpful to describe each file in the Supplemental file and explain where to find each file. Also a few readme files could be added to describe the figures in the `tune plots` folder and the `model fit tables` folder would be helpful.

### Summary assessment
-   fully and well documented

## Reproducibility

I was unable to render your manuscript as the issue was due to the code below 
```{r}
#| label: tbl-incfithb
#| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
readRDS(file = here("results", "model fit tables", "incfittable.rds")) 
```

If you run this it says the file can't be found. However it does exist but you forgot to add the `statistical_analysis` folder that lies in the `results folder`. I have gone ahead and made these changes shown in the code below
```{r}
#| label: tbl-incfithb
#| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
readRDS(file = here("results", "statistical_analysis", "model fit tables", "incfittable.rds")) 
```

```{r}
#| label: tbl-incfit
#| fig-cap: "Summary table for fitted linear model predicting incidence"
readRDS(file = here("results", "statistical_analysis","model fit tables", "incfittableFD.rds")) 
```

However when I attempted to render the file after updating the code, it looks like you left out the `statistical_analysis` folder for all of the results regarding the statistical analysis inside your manuscript. As a result, I have updated the code as an easy fix and it should all render now.

### Summary assessment
-   small parts not reproducible or required manual intervention

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

Your project was thorough as I felt that the main research question was answered and investigated in a structured and documented manner. I mentioned in the reveiw of your discussion/conclusion that some of the research questions were not fully answered but I do beleive you have the results to further discuss those questions.

### Summary assessment

-   strong level of thorougness

### 2023-05-01 LariscyReview.qmd

---
title: Project part five review
author: Leah Lariscy
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Tuberculosis Burden and Health Inequality Measures

Name of project author(s): Sara Benist

Name of project reviewer: Leah Lariscy

# Specific project content evaluation

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Your background was informative but concise, I would expand on it!

### Summary assessment (PICK ONE, DELETE THE OTHERS)

\* some contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

It is clear from the description what the aim of the study is and how it related to the data.

### Summary assessment

\* question/hypotheses fully clear

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The sources of the data were provided and characteristics of the data were described and it is clear how this data relates to the question.

### Summary assessment

\* source and overall structure of data well explained

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Wrangling steps were well-explained, however, some information like specific functions may be better in a supplementary file. I enjoyed your figures in the exploratory analysis, but some had labels that were difficult to read so those should be fixed.

### Summary assessment

\* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Analysis write-up was informative but short and the tables make it more visually cluttered and difficult to interpret. Describing the data and results with your woulds could be a better way to show your work. Explanations of your full analysis figures would also be beneficial as I am slightly lost on what is happening in your analysis.

### Summary assessment

\* defensible but not optimal analysis

## Presentation

How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Figures are pretty good, but the tables are cluttered and aren't quite publication quality. They are informative, but I think are more appropriate in a supplementary file.

### Summary assessment

\* results are presented ok, with room for improvement

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion and conclusion pretty much check all the boxes, however, I would remove the sentences mentioning meeting with the TA in the final manuscript.

### Summary assessment

\* minor parts wrong, missing or unclear

## Further comments

Overall I think you are trying to answer an interesting question and it seems like you are going about it the right way, but there needs to be more detail and clarity in the final manuscript. You may want to consider expanding upon the methods sections and analysis portions of the manuscript.

# Overall project content evaluation

Evaluate overall features of the project by filling in the sections below.

## Structure

Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Structure looks good!

### Summary assessment

\* well structured

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The documentation was well done, code had comments to describe the function and the I like how all your notes were in a .qmd and all code was knitted from a .R file.

### Summary assessment

\* fully and well documented

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I would add more detailed instruction on how to reproduce the manuscript, like which files are needed to reproduce all the figures.

### Summary assessment

\* fully reproducible without issues

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The hypothesis seemed to be pretty thoroughly addressed, and limitations of the study were addressed.

### Summary assessment

\* decent level of thoroughness

## Further comments

The project is really interesting and I can tell you put a lot of work into it but I think it needs a bit more flushing out of the details, especially in the analysis portion. The analysis wasn't easy to follow, so that is one area to work on.

### 2023-05-01 code/analysis_code/exploratory_analysis.qmd

@@
 # Setup
+
 The `exploratoryanalysis.r` script needs to be indicated as the file from which to pull code chunks from.
+
 ```{r, include=FALSE, cache=FALSE}
@@
 Next, load the needed packages.
+
 ```{r, packages, message = FALSE, warning = FALSE}
@@
 Then, load the data from the `processed_data` folder.
+
 ```{r, loaddata}
@@
 Similar to the cleaning code, I will separate the summary tables into groups of indicators.
+
 ```{r, exploratorytables1}
 ```
+
 The BCG coverage indicator has 4 dimensions and 12 subgroups and appears to have mostly data from smaller population sizes (most of the data has less than 720) and high BCG coverage (mean of 88.7%).
@@
 TB mortality is similar to the incidence variable in terms of subgroups, year, and population. The mortality outcome is also continuous, and the highest mortality rate is 140 deaths/100000 population.
+
 ```{r, exploratorytables3}
 ```
+
 Prevalence to notification ratio studies 2 subgroups within a dimension. There does not appear to be a trend toward shorter or longer length of time.
@@
 All data was collected from 2011 to 2019, and each group tends to have a relatively low percentage (means of 23.3%, 26.8%, and 19.9%).
+
 ```{r, exploratorytables5}
 ```
+
 Knowledge indicators of people reporting TB is spread through cough has similar dimensions and subgroups as the attitudes indicators.
@@
 ```
+
 The first plot show BCG coverage among under 1 year olds by economic status, education, place of residence, and sex.
@@
 Higher income seems to be correlated with lower catastrophic costs, and drug resistance has a much higher correlation with costs compared to non-drug resistant TB.
+
 ```{r, exploratoryfigures2}
 ```
+
 The third plot shows case detection rate by age and sex. Higher age groups tend to have higher detection rates (although the age group is quite large).
@@
 The forth plot shows drug resistance by sex. Neither group appears to have a higher percentage of drug resistance.
+
 ```{r, exploratoryfigures3}
 ```
+
 The 5a plot and 5b plots show TB incidence by country, separated by sex for the high burden countries and full data set.
@@
 There appears to be a small trend of urban areas having higher prevalence then rural.
+
 ```{r, exploratoryfigures4}
 ```
+
 The ninth plot shows the overall attitude towards keeping a family member's TB a secret by sex. Females tend to be correlated with a higher percentage compared to males.
@@
 ```
+
 The twelfth plot shows TB knowledge about spreading the disease through cough by sex. Males tended to have a slightly higher proportion compared to females.

### 2023-05-05 code/analysis_code/boosted_statistical_analysis.qmd
Final project commit

---
title: "Boosted Tree Statistical Analysis"
author: "Sara Benist"
date: "2023-04-06"
output: html_document
---

# Processing script with R script code

This Quarto file runs the statistical analysis code from the `booststatistical_analysis.r` script. The best way to view the output of the script is to render the qmd and view the html document. The processed data is used for model fitting.

*Note: this file takes approximately 15 minutes to render*

To start, the needed packages are loaded, and the `processeddata.rda` is loaded from the `processed_data` folder.

# Setup
The `statistical_analysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, cache=FALSE}
knitr::read_chunk('booststatistical_analysis.R')
```

# Load packages and data
The packages and data are loaded.
```{r, packages, message = FALSE, warning = FALSE}
```

```{r, loaddata}
```

# Model fitting
In addition to the decision tree modelling found in `statistical_analysis.qmd` file, I wanted to also look at boosted tree models. First, we will split the data like in the decision tree models.

```{r, splitdata}
```
I will be tuning boosted tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the full data set. To make it more clear when saving the resulting tables and plots, the other burden country boosted tree results will be labeled `obb` and high burden as `hbb`.

## Boosted tree model for TB Incidence
```{r, obbincidencemodel}
```
The best boosted tree model for TB incidence using the other burden countries has a depth of 15 with an RMSE of 48 (SE = 4.81). This is slightly better than the decision tree model (RMSE of 51.6). The fitted model to the training data has a RMSE of 49.0, and since the model performs better than the  decision tree, we can fit the test data as well. Using the testing data, the model had a performance of RMSE = 37.1. This would indicate the model would perform similar on new data (compared to the training data). The residual plot shows clustering in the low prediction values with high residuals for 0-100 predictions.

```{r, hbbincidencemodel}
```
The best boosted tree model for TB incidence using the high burden countries has a depth of 8 with an RMSE of 134 (SE = 8). This is also a low level of performance in predicting TB incidence but slightly improved from the decision trees (RMSE of 159. The fitted model to the training data has a RMSE of 136 which was much higher compared to the full dataset. Using the testing data, the model had a performance of RMSE = 92.4. The residual plot shows clustering for low values but very low residuals for all values.

## Boosted tree model for TB Mortality
```{r, obbmortalitymodel}
```
The best boosted tree model for TB mortality using the other burden countries has a depth of 1 with an RMSE of 9.08 (SE = 1.49). The fitted model to the training data has a RMSE of 8.66, and using the testing data, the model had a performance of RMSE = 6.34. The residual plot shows clustering in low values and moderate spread.

```{r, hbbmortalitymodel}
```
The best boosted tree model for TB mortality using the high burden countries has a depth of 15 with an RMSE of 18.2 (SE = 1.74). The fitted model to the training data has a RMSE of 18.5. Using the testing data, the model had a performance of RMSE = 11.1. The residual plot shows a uniform spread with low low residual values.

## Boosted tree model for TB Prevalence
```{r, obbprevalencemodel}
```
The best boosted tree model for TB prevalence using the other burden countries has a depth of 8 with an RMSE of 323 (SE = 130). This is a very low level of performance in predicting TB prevalence with high uncertainty. The fitted model to the training data has a RMSE of 450 which is much worse compared to the decision tree models. For comparison since the other boosted tree models all preformed better than the decision trees, I will still fit the testing data to the boosted tree models. Using the testing data, the model had a performance of RMSE = 257. The residual plot shows only positive residuals (indicating the actual values were far greater than the predictions).

```{r, hbbprevalencemodel}
```
The best boosted tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 322 (SE = 58.9). This is a very low level of performance in predicting TB incidence but has less uncertainty than the other burden model. The fitted model to the training data has a RMSE of 351, but using the testing data, the model had a performance of RMSE = 448 which is the worst performance of the models. The  residual plot still is mostly high value, positive residuals but is slightly better than the other model.

## Summary
```{r, compare}
```
The models were overall not subpar predictors of TB outcomes. TB prevalence was significantly different from the incidence and mortality models, possibly due to using different predictor variables or the limited availability of data points. The mortality models performed best based on RMSE values, and the other burden incidence and mortality models performed better than the high burden countries.

### 2023-05-05 code/analysis_code/statistical_analysis.qmd
Part 5 of project
cleaned up manuscript, removed extra text, fixed graphs/plots
Final project commit

--- 2023-03-17 code/analysis_code/statistical_analysis.qmd
@@
 ---
-title: "Statistical Analysis"
+title: "Decision Tree Statistical Analysis"
 author: "Sara Benist"
-date: "2023-03-17"
+date: "2023-04-06"
 output: html_document
@@
 
-This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The processed data is used for model fitting.
+This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script.
+The best way to view the output of the script is to render the qmd and view the html document.
+The processed data is used for model fitting.
+
+*Note: this file takes approximately 10 minutes to render*
 
@@
 ```
+
+# Basic modelling
+For some basic modeling, I will be fitting linear models to the TB outcomes with the subgroups as predictors.
+Subgroups are the main difference between populations other than country which is why I chose these as the bivariate models.
+This will offer another view of the data.
+The prediction model also offers a way to explore the data by submitting a sample prediction of TB incidence for females and males in Central African Republic.
+```{r, basicmodels}
+```
+The models indicate that subgroup has a significant influence on the outcomes since the estimates for the subgroup (being Male primarily or living in rural areas) increases the incidence, mortality, or prevalence.
+This is expected from the exploratory analysis.
+
+The biggest issue with the linear modeling is the assumption that the relationship between predictors and outcomes is linear, and the model cannot be tuned and replicated.
+
 # Model fitting
 Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set.
-First, we will focus on linear models of the outcomes TB incidence, TB mortality, and TB prevalence.
-I will be fitting two models using the full data set and only the high burden countries.
+First, we will split the data into testing and training sets and set up the cross-validation folds.
+The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.
 
-## Linear model fitting for TB Incidence
+```{r, splitdata}
+```
 
-```{r, fitmodel1}
+I will be tuning decision tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the other burden set.
+In the `boosted_statistical_analysis.qmd` file, a boosted tree model for each outcome is created for comparison.
+In order to be clear when saving the resulting tables and figures, the other burden outcomes from the decision tree model are labeled as `obdt` and the high burden as `hbdt`.
+The testing data will be fit to the best performing models.
+
+## Decision tree model for TB Incidence
+```{r, obdtincidencemodel}
 ```
-The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
-The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`.
-Since many of the countries in the full data set have very low incidence, the coefficients for the intercept and subgroup variable are much lower.
-The model has more data points with low values.
+The best tree model for TB incidence using the other burden countries has a depth of 4 with an RMSE of 51.6.
+This is a very low level of performance in predicting TB incidence (ideally would be closer to 1).
+The fitted model to the training data has a RMSE of 29.8.
 
-### Prediction model for TB Incidence
-We can also try to predict the approximate TB incidence for a group. For this example, we will include subgroup and setting as predictors.  
-```{r, pred1}
+```{r, hbdtincidencemodel}
 ```
-With the prediction model including setting, there are a lot of variables in the model.
-The prediction would have a 1 for the country of interest and 0 for all other setting variables.
-For example, the prediction for a male in Central African Republic would appear as `TB incidence = 126(1)+189(1)`.
+The best tree model for TB incidence using the high burden countries has a depth of 4 with an RMSE of 159.
+This is also a low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 81.3 which was much higher compared to the full dataset.
 
-As an example, I will predict the incidence rate for males and females in India.
-The code for prediction modeling is based off of the `tidymodels` guidance.
-The plot shows the incidence rate and confidence intervals for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) in India.
-This can be repeated with the high burden countries.
+## Decision tree model for TB Mortality
+```{r, obdtmortalitymodel}
+```
+The best tree model for TB mortality using the other burden has a depth of 8 with an RMSE of 8.85. The fitted model to the training data has a RMSE of 7.21.
 
-## Linear model fitting for TB Mortality
+```{r, hbdtmortalitymodel}
+```
+The best tree model for TB mortality using the high burden countries has a depth of 4 with an RMSE of 20.4. The fitted model to the training data has a RMSE of 10.3.
 
-```{r, fitmodel2}
+## Decision tree model for TB Prevalence
+
+```{r, obdtprevalencemodel}
 ```
-For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
+The best tree model for TB prevalence using the other burden countries has a depth of 1 with an RMSE of 294.
+This is a very low level of performance in predicting TB prevalence.
+The fitted model to the training data has a RMSE of 240.
 
-## Linear model fitting for TB Prevalence
+```{r, hbdtprevalencemodel}
+```
+The best tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 322.
+This is a very low level of performance in predicting TB incidence.
+The fitted model to the training data has a RMSE of 254.
 
-```{r, fitmodel3}
+The models were overall not a good predictor of TB outcomes.
+This is possibly due to the large variance in the outcomes as well as the summarizing required to de-aggregate the data and predict outcomes.
+Country was overwhelmingly the most important predictor.
+
+## Summarise models
+Overall, the models were not great predictors of TB outcomes. The mortality performed the best, and all high burden models had lower RMSE values by about half of the other burden models. 
+In the `boosted_statistical_analyisis.qmd` file, I created boosted tree models as a comparison. The boosted tree models should perform better than a decision tree model. 
+```{r, summary}
 ```
-For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0.
-The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.
-The difference between the high burden and full data models is lower than the other outcomes, most likely due to the long duration of TB or less data points available with the full data model.
-
-## Linear model fitting for BCG Coverage
-Now we can look at some of the other indicators in the data set. I will only be looking at the indicators that had variation between subgroups. 
-```{r, fitmodel4}
-```
-The BCG coverage has 12 variables that produce a linear equation of `BCG coverage % = 88.4 + 0.363(Male) -10.6(no education) + 0.169(primary education) + 6.66(higher education) - 6.46(quintile 1) - 1.6(quintile 2) +1.06(quintile 3) + 4.50(quintile 4) + 7.01(quintile 5) - 2.12(rural) + 5.10(Urban)`.
-The full model has an equation of `BCG coverage % = 88.3 + 0.245(Male) -9.44(no education) + 0.958(primary education) + 6.42(higher education) - 6.40(quintile 1) - 1.68(quintile 2) + 0.773(quintile 3) + 3.84(quintile 4) + 6.41(quintile 5) - 2.08(rural) + 4.75(Urban)`.
-The two models are overall pretty similar and can be used in the same way as the prediction model (entering in 1 or 0 depending on categorization).
-
-## Linear model fitting for Catastrophic costs
-
-```{r, fitmodel5}
-```
-The high burden model has an equation of `% catastrophic costs = 87.3 - 40.4(drug-suseptible TB) - 18.6(quintile 1) - 38.8(quintile 2) - 44.6(quintile 3) - 47.4(quintile 4) - 58.5(quintile 5)`.
-The full model data has an equation of `% catastrophic costs = 83.5 - 35.1(drug-suseptible TB) - 11.1(quintile 1) - 30.9(quintile 2) - 38.4(quintile 3) - 41.4(quintile 4) - 54.9(quintile 5)`.
-## Linear model fitting for case detection rate
-
-```{r, fitmodel6}
-```
-For high burden countries, the equation is `case detection rate % = 37.9 + 19.1(15+ or older) + 16.7(female) + 16.4(male)`.
-The full data model has an equation of `case detection rate % = 43.8 + 18.9(15+ or older) + 15.3(female) + 16.9(male)`.
-
-## Linear model fitting for prevalence to notification ratio
-
-```{r, fitmodel7}
-```
-The high burden model has an equation of `prevalence to notification ratio in years = 1.73 + 0.664(male)`, and the full model has an equation of `prevalence to notification ratio in years = 1.73 + 0.701(male)`.
-
-## Linear model fitting for attitudes
-
-```{r, fitmodel8}
-```
-The attitude question regarding TB was about the desire to keep a diagnosis hidden.
-Since males are more likely to be diagnosed with TB, I focused on the male attitude and knowledge.
-For high burden countries, the equation is `% wanting to keep TB secret = 27.6 - 5.27(20-34 years) - 8.21(35-49 years) - 3.53(no education) - 2.71(primary education) - 7.23(higher education) - 5.01(quintile 1) - 5.23(quintile 2) - 4.97(quintile 3) - 4.38(quintile 4) - 5.86(quintile 5) - 4.91(rural) - 6.27(urban)`.
-For the full model, the equation is `% wanting to keep TB secret = 24.7 - 4.94(20-34 years) - 8.14(35-49 years) - 4.32(no education) - 3.08(primary education) - 6.41(higher education) - 4.44(quintile 1) - 5.24(quintile 2) - 5.18(quintile 3) - 4.59(quintile 4) - 6.20(quintile 5) - 4.66(rural) - 5.99(urban)`.
-
-## Linear model fitting for knowledge
-
-```{r, fitmodel9}
-```
-The knowledge question asked about if TB is spread through cough. We will again focus on male knowledge.
-
-The high burden countries had an equation of `% responding TB is spread through cough = 65.9 + 4.83(20-34 years) + 6.19(35-49 years) - 9.91(no education) - 2.81(primary education) + 12.6(higher education) - 6.01(quintile 1) - 0.464(quintile 2) + 2.32(quintile 3) + 7.50(quintile 4) + 14.6(quintile 5) + 0.419(rural) + 11.8(urban)`.
-The full model had an equation of `% responding TB is spread through cough = 64.6 + 6.17(20-34 years) + 7.29(35-49 years) - 8.55(no education) - 1.66(primary education) + 12.8(higher education) - 3.88(quintile 1) + 0.028(quintile 2) + 3.03(quintile 3) + 8.18(quintile 4) + 15.8(quintile 5) + 1.28(rural) + 12.8(urban)`.
-
-I would like to try and create a dataset using the high burden countries in order to predict the main outcomes of TB incidence and mortality from the indicators (BCG Coverage, knowledge, attitudes, etc).
-Right now, the data is not set up to complete this analysis, but I will try to find a way to do this.

### 2023-05-05 code/processing_code/processingfile.qmd
Final project commit

@@
 I will subset these countries in the `highburden` object.
+The countries with low or mid level burden will be sorted into `otherburden`.
 
@@
 
-The `wide_data` had 7473 rows and 20 columns. The `highburden` object has 3761 rows and 20 columns.
+The `wide_data` had 7473 rows and 20 columns with 194 countries.
+The `highburden` object has 3761 rows and 20 columns with 49 countries.
+The `otherburden` object has 3712 rows and 20 columns with 145 countries.
 
@@
 
+# Modelling setup
+
+The final analysis will be predicting tuberculosis outcomes from the indicators.
+Due to the nature of the data, each indicator must be summarized for males and females (aggregate data can measure males or someone in the lowest economic quintile but cannot predict the indicator measure for males in the lowest economic quintile).
+To set up the data for modelling, the data is summarized by sex and indicator for each country in the high burden and other burden data sets.
+The incidence and mortality outcomes consist of indicators with sex as a dimension, and the prevalence outcome consists of rural or urban subgroups.
+
+```{r, modelsetup}
+```
+
 # Save data

### 2023-05-05 products/manuscript/Manuscript.qmd
Final project commit

@@
 library(knitr)
-library(tidyverse)
 ```
@@
 The World Health Organization collects annual data on tuberculosis outcomes through the Global Tuberculosis Programme with the goal of reducing TB incidence worldwide.
-The further explore the reasons behind the health inequity surrounding tuberculosis, this analysis explores the health equity indicators collected by WHO and attempts to predict tuberculosis outcomes.
+To further explore the reasons behind the health inequity surrounding tuberculosis, this analysis explores the health equity indicators collected by WHO and attempts to predict tuberculosis outcomes.
 
@@
 The data set was also subset into high burden countries to explore difference between high burden countries and all countries.
-Linear regression and decision tree modelling using cross-validation tuning were conducted with predict tuberculosis outcomes and level of equity indicators.
+Linear regression along with decision tree and boosted tree modelling using cross-validation tuning were conducted with predict tuberculosis outcomes and level of equity indicators.
 
-In 2020, Lesotho had the highest TB incidence, but with the inclusion of mortality rates, Central African Republic had the highest case fatality rate with a high incidence and mortality.
 Males have a higher proportion of cases compared to females, and level of income and education had the largest disparity between the highest and lowest levels.
+The prediction models did not perform well in predicting tuberculosis outcomes, most likely due to the aggregate nature of the data.
 More research needs to be conducted to measure the direct relationship between the health equity indicators and tuberculosis outcomes.
@@
 ## Questions/Hypotheses to be addressed
-*update depending on ability to bind indicators with outcomes; discuss with Zane next week*
 
-Research question: Does high disparities between inequality measures show high correlation to TB incidence, prevalence, and mortality of a country?
+Research question: How does disparities within the health equity impact the indicator estimates?
+Does high disparities between inequality measures show high correlation to TB incidence, prevalence, and mortality of a country?
 How do these relationships differ for high burden countries compared to all countries?
@@
 The overall outcome I would study is TB mortality since effective health programs ideally reduce disease-specific mortality.
-By the end of my analysis, I would like to be able to identify populations that could be a focus of TB health improvement programs.
-Other outcomes I would like to explore include regional differences in drug-resistant TB and the attitudes and perceptions for high burden areas compared to medium or low burden regions.
+The intention of this project is to identify populations that could be a focus of TB health improvement programs.
 
@@
 The data cleaning will consist of widening the data set so each TB indicator is placed in a column in order to explore the indicators separately.
-The data set will also be split into a full data set containing all countries (wide_data) and high burden countries (highburden).
+The data set will also be split into two data sets containing low/mid burden countries (otherburden) and high burden countries (highburden).
 The data will then be summarized and described using exploratory analysis approaches such as graphs, plots, and tables.
-The statistical analysis will use cross-validation and decision tree modeling to model TB outcomes based on the available predictor variables.
-A similar approach will be used for the health equity indicators as outcomes.
+The statistical analysis will use decision tree and boosted tree modeling to model TB outcomes based on the available predictor variables.
+In order to complete the modelling, the data sets must be summarized so each country has only two values for each available indicator, one value for male and one for female.
+This is required because the data is aggregated, but much information on the disparities within indicators is lost.
 
@@
 
-For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden`.
+For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden` and all other countries into `otherburden`.
+In addition, I set up the modeling data into `hbmodel` and `obmodel` which contain indicators with sex as a dimension and `hbmodelprev` and `obmodelprev` for those with place of residence.
 I also checked each indicator using `filter()`, `select()`, and `summary()` functions.
@@
 
-For the statistical analysis of the data, I created decision tree models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator and tuned the models using cross-validation folds.
-The data was initially split into test/train data sets to better evaluate model performance.
-This method was repeated for the `wide_data` and `highburden` countries.
+The exploratory analysis consisted of describing the disparities within each indicator and for the TB outcomes.
+This process assisted in understanding the data and framing the data for model exploration.
+The full exploratory analysis can be found in the `exploratory_analysis.qmd` file and the associated r script.
 
-I would still like to try to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes.
-This will most likely be limited to the high burden countries.
-I am meeting with Zane next week to discuss.
+The basic statistical analysis consisted of bivariate analyses of the TB outcomes based on subgroups identified in the exploratory analysis.
+For the model analysis of the data, I initially created decision tree models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the indicator estimates for females and males and tuned the models using cross-validation folds.
+The data was split into test/train data sets to better evaluate model performance.
+This method was repeated for the `otherburden` and `highburden` countries.
+The full statistical analysis for decision tree modeling can be found in the `statistical_analysis.qmd` file.
 
+As a means of comparison, I also created boosted tree models for the TB outcomes following the same general methods as the decision tree modelling.
+The full analysis can be found in the `boosted_statistical_analysis.qmd` file.
 {{< pagebreak >}}
@@
 
-After processing the data, the `wide_data` and `highburden` datasets were described by exploring each health equity indicators and tuberculosis outcome.
-@fig-indicators shows the list of indicators, and the codebook can be found in the README.md file.
+After processing the data, the `otherburden` and `highburden` datasets were described by exploring each health equity indicators and tuberculosis outcome.
+The table below shows the list of indicators, and the codebook can be found in the README.md file.
 
@@
 The summary tables are stored in the `results` folder.
+
 The population for most variables were relatively low and skewed to the right.
 The TB incidence and mortality indicators also showed a skewed distribution.
-The highest incidence rate was 908 cases/100,000 shown in @fig-incsum, and highest mortality rate was 140 deaths/100,000 shown in @fig-mortsum.
-```{r}
-#| label: fig-incsum
-#| fig-cap: "Tuberculosis incidence summary"
-readRDS(file = here("results", "exploratory_analysis", "summary tables", "incsummarytable.rds"))
-```
-
-```{r}
-#| label: fig-mortsum
-#| fig-cap: "Tuberculosis mortality summary"
-readRDS(file = here("results", "exploratory_analysis", "summary tables", "mortsummarytable.rds"))
-```
+The highest incidence rate was 908 cases/100,000 , and highest mortality rate was 140 deaths/100,000.
 
@@
 ```
+
 @fig-incexplore shows the tuberculosis incidence for each country stratified by sex.
@@
 @fig-mortexplore shows the TB mortality for high burden countries. Compared with the TB incidence values, Central African Republic has high TB incidence and mortality.
+
 ```{r}
@@
 @fig-prevexplore shows tuberculosis prevalence by country. There may be a slight trend of urban areas having higher prevalence compared to rural areas.
+
 ```{r}
@@
 The economic status and education levels showed an approximate difference of 10-15% vaccination coverage between the highest and lowest levels.
+
 ```{r}
@@
 ```
+
 Experiencing catastrophic costs due to TB was most heavily present in the lowest economic quintile and was highly associated with drug resistant TB with approximately 85% of people with drug-resistant strains experiencing high costs.
+
 ```{r}
@@
 Males on average have 2.4 years between contracting tuberculosis and receiving a diagnosis compared to 1.7 years for females.
+
 ```{r}
@@
 
-The final indicator that showed the highest disparity is the male and female knowledge of TB shown in @fig-knowledge.
+The final indicator that showed the highest disparity is the male and female knowledge of TB shown in Figure 8.
 Both groups show similar disparities within the economic status, education level, and place of residence subgroups.
@@
 Similar trends occur for th education and place of residence subgroups.
+
 ```{r}
@@
 Most indicators showed the greatest disparities due to economic status and education level.
-Please render the `exploratory_analysis.qmd` file for the full exploratory analysis exploring the other health equity indicators.
 
 ## Basic statistical analysis
-*update depending on binding outcomes and predictors: will probably do linear model predictions from TB outcomes from one indicator if possible.*
 
-The linear models summary for TB incidence are listed below as @tbl-incfithb and @tbl-incfit.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
-The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`.
+The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1.
+The other burden model has the linear equation of `TB incidence = 44.9 + 21.5(Male)`.
 Males having higher TB incidence is congruent with the exploratory analysis findings.
-The full model has smaller coefficients compared to `highburden` model.
+The `otherburden` model has smaller coefficients compared to `highburden` model.
 
 ```{r}
-#| label: tbl-incfithb
-#| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
 readRDS(file = here("results", "statistical_analysis", "model fit tables", "incfittable.rds")) 
@@
 
+"Summary table for fitted linear model predicting incidence for high burden countries"
+
 ```{r}
-#| label: tbl-incfit
-#| fig-cap: "Summary table for fitted linear model predicting incidence"
 readRDS(file = here("results", "statistical_analysis","model fit tables", "incfittableFD.rds")) 
@@
 
-The fitted model summaries below are the TB mortality for high burden countries (@tbl-mortfithb) and full data (@tbl-mortfit).
-For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.
+"Summary table for fitted linear model predicting incidence"
+
+The fitted model summaries below are the TB mortality for high burden countries and other burden data.
+For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the other burden model, the linear equation is `TB mortality = 5.36 + 3.56(Male)`.
 
 ```{r}
-#| label: tbl-mortfithb
-#| fig-cap: "Summary table for fitted linear model predicting mortality for high burden countries"
 readRDS(file = here("results", "statistical_analysis", "model fit tables", "mortfittable.rds"))
 ```
+
+"Summary table for fitted linear model predicting mortality for high burden countries"
+
 ```{r}
-#| label: tbl-mortfit
-#| fig-cap: "Summary table for fitted linear model predicting mortality"
 readRDS(file = here("results","statistical_analysis", "model fit tables", "mortfittableFD.rds"))
@@
 
-The last outcome model I fitted was TB prevalence for high burden countries (@tbl-prevfithb) and the full data (@tbl-prevfit).
+"Summary table for fitted linear model predicting mortality"
+
+The last outcome model I fitted was TB prevalence for high burden countries and the other burden data.
 For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0.
-The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`.
+The full data model has an equation of `TB Prevalence = 444 - 71.4(Urban)`.
+Interestingly, the place of residence had an opposite effect on TB prevalence between the high burden and other burden models.
 
 ```{r}
-#| label: tbl-prevfithb
-#| fig-cap: "Summary table for fitted linear model predicting prevalence for high burden countries"
 readRDS(file = here("results","statistical_analysis", "model fit tables", "prevfittable.rds"))
 ```
+
+"Summary table for fitted linear model predicting prevalence for high burden countries"
+
 ```{r}
-#| label: tbl-prevfit
-#| fig-cap: "Summary table for fitted linear model predicting prevalence"
 readRDS(file = here("results", "statistical_analysis", "model fit tables", "prevfittableFD.rds")) 
 ```
+
+"Summary table for fitted linear model predicting prevalence"
 
@@
 
-For the majority of the statistical analysis, the tuberculosis outcomes and the health equity indicators with high disparity between subgroups (as identified during the exploratory analysis) were modeled and tuned using the decision tree model and cross-validation methods from the `tidymodels` guide.
-The models were all used the training data to tune the cost_complexity and tree_depth parameters to find the model with the lowest value of RMSE.
+For the majority of the statistical analysis, the tuberculosis outcomes were modeled using the decision tree and boosted tree models and tuned with cross-validation methods from the `tidymodels` guide.
+The decision tree models all used the training data to tune the cost_complexity and tree_depth parameters to find the model with the lowest value of RMSE.
+The boosted tree models used tree_depth, min_n, and trees as tuning parameters.
 
-The @tbl-treeinc displays an example of the model tuning for decision tree model predicting incidence, and @tbl-treeinchb shows the same method for high burden countries.
-Please refer to the `statistical_analysis.qmd` for the full code and results on modelling.
+The @fig-treeinc displays an example of the model tuning for decision tree model predicting incidence for other burden countries, and @fig-treeinchb shows the same method for high burden countries.
+Please refer to the `statistical_analysis.qmd` for the full code and results on decision tree modelling.
+
 ```{r}
-#| label: tbl-treeinc
+#| label: fig-treeinc
 #| fig-cap: "Summary table for fitted tree model predicting incidence"
-readRDS(file = here("results", "statistical_analysis","tune plots", "wdinctreeplot.rds")) 
+readRDS(file = here("results", "statistical_analysis","tune plots", "obdtinctreeplot.rds")) 
 ```
@@
 ```{r}
-#| label: tbl-treeinchb
+#| label: fig-treeinchb
 #| fig-cap: "Summary table for fitted linear model predicting incidence for high burden countries"
-readRDS(file = here("results", "statistical_analysis","tune plots", "hbinctreeplot.rds"))
+readRDS(file = here("results", "statistical_analysis","tune plots", "hbdtinctreeplot.rds"))
 ```
 
-A summary of the model performances are shown below predicting TB outcomes and healthy equity indicators by subgroup and setting.
-The full data set models are marked with 'wd' and the high burden models are marked as 'hb'.
+A summary of the model performances are shown below predicting TB outcomes.
+The `otherburden` models are marked with 'obdt' and the high burden models are marked as 'hbdt'.
 The outcomes being modeled are abbreviated in the `model` column.
-The best performing model predicted prevalence to notification ratio for the full data set (RMSE = 0.971).
+The best performing model predicted mortality for other burden countries (RMSE = 10.9).
 The TB outcome models all performed poorly in predicting TB incidence, mortality, and prevalence.
-This is most likely due to the only significant predictor being country.
+The `otherburden` models performed better than the high burden models for incidence and mortality, and prevalence models performed worse overall.
+
 ```{r}
-#| label: tbl-rmsesum
-#| fig-cap: "Summary table of RMSE values for tree models"
-readRDS(file = here("results", "summaryrmse.rds"))
+readRDS(file = here("results", "statistical_analysis", "dtsummaryrmse.rds"))
 ```
 
-The second research question was to explore the differences between the full data set and high burden countries.
-I hypothesized that the high burden models would have a higher performance because these countries have the highest concentration of TB cases.
-However, the high burden models consistently preformed worse than the full data models.
-This is most likely due to the lower amount of available data.
+"Summary table of RMSE values for decision tree models"
+
+The boosted tree models performed better than the decision tree models for the most part.
+The `model` section indicates the model being created.
+The mortality models performed best (RMSE of 8.66 and 18.5), and the prevalence models preformed significantly worse compared to the other models (RMSE of 450 and 350.7).
+
+```{r}
+readRDS(file = here("results", "statistical_analysis", "btsummaryrmse.rds"))
+```
+
+"Summary table of RMSE values for boosted tree models"
+
 {{< pagebreak >}}
@@
 The objective of this project was to compare disparities in the health equity indicators with tuberculosis outcomes to determine if larger disparities are associated with tuberculosis burden.
-Due to the formatting of the data, the relationship could not be directly compared.
-Instead, each equity indicator was individually analyzed, and the tuberculosis outcomes were modeled using the countries and subgroups.
+Due to the formatting of the data, the relationship could not be directly compared since the data is aggregated (the BCG coverage can be measured for low income individuals and females but cannot measure a low income female).
+Instead, each equity indicator was individually analyzed through exploratory tables and graphs, and the tuberculosis outcomes were modeled using summarised values of the indicators with sex or place of residence as dimensions.
 
@@
 The incidence and mortality models were consistent with the exploratory analysis since being male was the main predictor of the model.
-The high burden countries have higher levels of TB outcomes compared to the full data set.
-However, these models did not take into account the country as a predictor, so the full analysis used a decision tree model to include this as a predictor.
+The high burden countries have higher levels of TB outcomes compared to the other burden countries which is consistent with the existing literature.
+However, these models did not take into account other variables as a predictor, so the full analysis used a decision tree model and boosted tree model to predict outcomes.
 
-The full analysis consisted of a series of decision tree models that predicted TB outcomes and equity indicators by subgroups.
-The high burden models were predicted to perform better than the full set models, but this hypothesis did not hold true.
-All tuberculosis outcome models did not perform well with the incidence, mortality, and prevalence models for the high burden countries having RMSE values of 163, 21.6, and 184 respectively.
-The prevalence model was the only outcome in which the high burden model performed better than the full model.
+The full analysis initially consisted of a series of decision tree models that predicted TB outcomes from summarized health equity indicators.
+The high burden models were predicted to perform better than the other burden models, but this hypothesis did not hold true.
+No tuberculosis outcome models performed well, but the boosted tree models performed better compared to the decision tree models for TB incidence and mortality with high burden RMSE values of 135.5 and 18.5 and other burden RMSE values of 51 and 8.7 respectively.
+The prevalence model was the only outcome in which the high burden model performed better than the other burden model.
 
-Of the healthy equity indicators, the prevalence to notification ratio, attitudes toward TB, and BCG vaccination coverage models were the top performing models.
+The prevalence model preformed significantly different from the incidence and mortality models.
+One hypothesis for the difference is the less available data for TB prevalence compared to the other outcomes or the difference in variables being used as predictors since the TB prevalence used different subgroups.
 
-For most decision tree models, country was the most significant predictor.
-This indicates that public health programs should personalize their efforts to eliminate tuberculosis using country/population specific factors.
-
-After discussing with Zane, the goal would be to be able to use health equity indicator level (ex: BCG vaccine coverage) to predict TB outcome.
-With this information, you would be able to model how changes would improve TB outcome (if you raise the BCG vaccine coverage to at least 80%, the predicted TB incidence would lower by ##).
+While these models are do not perform well in predicting TB outcomes, the exploratory analysis showed several populations that could be the focus of public health programs in improving health disparities.
+Education level and economic status showed the greatest disparity within health indicators, so public health programs can be designed to focus on improving or addressing the health equity indicators, such as hosting BCG vaccine efforts in low income populations, offering tuberculosis education courses for populations with limited educational resources, or encouraging males to be tested for TB often.
 
@@
 This project had several limitations associated with the analysis.
-Due to the collection years differing between indicator variables and TB outcomes, I am not able to predict tuberculosis incidence, mortality, and prevalence from the health inequity indicators.
-In order to complete this analysis as my research question designed, I would be losing valuable data on the subgroups by having to summarize the overall indicators by country.
-I will be discussing this with Zane and either adjust my research question or reformat my data.
-Another limitation of the analysis is the difficulty interpreting the decision tree models.
-Since country was the most significant predictor for the models, it was difficult to visualize and express the model.
+Due to aggregate nature of the data and the collection years differing between indicator variables and TB outcomes, I am not able to predict tuberculosis incidence, mortality, and prevalence from the health inequity indicators directly.
+This aggregation is used to protect the identity of the program participants, so I do not think de-aggregated information would be publicly available.
+In order to complete this analysis as my research question required, I had to summarize the indicator values which lost almost all the disparities measured within the data.
+I believe this reduced the performance of my models because the predictors would have been more similar to each other that did not coordinate with the TB outcomes.
+Another limitation of the analysis is the difficulty interpreting the decision tree and boosted tree models.
+It was difficult to visualize and express the model due to the large number of country predictors.
 The final limitation is that the data did not capture all factors that contribute to the spread of tuberculosis such as population density, access to treatment regimens, consistency of adherence, and other aspects that can affect one's susceptibility to contracting or spreading the disease.
@@
 ## Conclusions
+
 This analysis explores tuberculosis-related disparities between levels of income, education, sex, and other health equity topics.
 TB incidence and mortality is most closely associated with the country.
-There is no direct comparison between level of disparity in equity indicator and tuberculosis incidence, prevalence, or mortality.
+There is no direct comparison between level of disparity in equity indicator and tuberculosis incidence, prevalence, or mortality due to the aggregated data set, so the prediction models poorly performed.
 Continued analysis is required to address the gap in data preventing a direct measurement of health disparity on tuberculosis outcomes.

### 2023-05-05 review_Jacob_Matta.qmd
Final project commit

@@
 title: Project Review Template 
-author: 
+author: Jacob Matta
 date: "`r file.mtime(knitr::current_input())`"
@@
 I was unable to render your manuscript as the issue was due to the code below
-```{r}
+
 #| label: tbl-incfithb
@@
 readRDS(file = here("results", "model fit tables", "incfittable.rds"))
-```
 
@@
 I have gone ahead and made these changes shown in the code below
-```{r}
+
 #| label: tbl-incfithb
@@
 readRDS(file = here("results", "statistical_analysis", "model fit tables", "incfittable.rds"))
-```
 
-```{r}
 #| label: tbl-incfit
@@
 readRDS(file = here("results", "statistical_analysis","model fit tables", "incfittableFD.rds"))
-```
 

### 2023-05-05 code/analysis_code/booststatistical_analysis.R

@@ -1,0 +1,496 @@
+###############################
+# analysis script
+#
+#this script loads the processed, cleaned data, does a simple analysis
+#and saves the results to the results folder
+
+## ---- packages --------
+#load needed packages
+library(here) 
+library(tidymodels)
+library(tidyverse)
+library(ggplot2)
+library(bonsai)
+
+## ---- loaddata --------
+#path to data
+data_location <- here::here("data","processed_data","processeddata.rda")
+
+#load data. 
+load(data_location)
+
+## ---- splitdata --------
+#split data into training and test sets for wide_data
+obbdatasplit <- initial_split(obmodel, prop = 3/4)
+obbtest_data <- testing(obbdatasplit)
+obbtrain_data <- training(obbdatasplit)
+
+#set seed for reproducibility
+set.seed(626)
+
+#create cross-validation folds
+obbfolds <- vfold_cv(obbtrain_data, v = 5)
+
+#split data into training and test sets for highburden
+hbbdatasplit <- initial_split(hbmodel, prop = 3/4)
+hbbtest_data <- testing(hbbdatasplit)
+hbbtrain_data <- training(hbbdatasplit)
+
+#set seed for reproducibility
+set.seed(626)
+
+#create cross-validation folds
+hbbfolds <- vfold_cv(hbbtrain_data, v = 5)
+
+#split data into training and test sets for other burden prevalence
+obbprevsplit <- initial_split(obmodelprev, prop = 3/4)
+obbprevtest_data <- testing(obbprevsplit)
+obbprevtrain_data <- training(obbprevsplit)
+
+#set seed for reproducibility 
+set.seed(626)
+
+#create cross-validation folds
+obbprevfolds <- vfold_cv(obbprevtrain_data, v = 5)
+
+#split data into training and test sets for high burden prevalence
+hbbprevsplit <- initial_split(hbmodelprev, prop = 3/4)
+hbbprevtest_data <- testing(hbbprevsplit)
+hbbprevtrain_data <- training(hbbprevsplit)
+
+#set seed for reproducibility 
+set.seed(626)
+
+#create cross-validation folds
+hbbprevfolds <- vfold_cv(hbbprevtrain_data, v = 5)
+
+## ---- obbincidencemodel ---------
+#recipe predicting TB incidence
+increcipe <- recipe(`TB incidence (new infections per 100 000 population)` ~.,
[…]

### 2023-05-05 code/analysis_code/exploratoryanalysis.R

@@ -13,6 +13,7 @@ library(skimr)
 library(ggplot2) #for plots/figures
 library(forcats) #for factoring
 library(scales)
+library(tidyr)
 
 
 ## ---- loaddata --------
@@ -23,89 +24,101 @@ load(data_location)
 
 ## ---- exploratorytables1 --------
 #bcg coverage summary and save to file location
-summary_bcg <- skimr::skim(wide_data$`BCG immunization coverage among one-year-olds (%)`)
+summary_bcg <- skimr::skim(otherburden$`BCG immunization coverage among one-year-olds (%)`)
 print(summary_bcg)
 bcgsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "bcgsummarytable.rds")
 saveRDS(summary_bcg, file = bcgsummarytable_file)
 
 #catastrophic cost summary and save to file location
-summary_catacost <- skimr::skim(wide_data$`Families affected by TB facing catastrophic costs due to TB (%)`)
+summary_catacost <- skimr::skim(otherburden$`Families affected by TB facing catastrophic costs due to TB (%)`)
 print(summary_catacost)
 catacostsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "catacostsummarytable.rds")
 saveRDS(summary_catacost, file = catacostsummarytable_file)
 
 #case detection rate summary and save to file location
-summary_cdr <- skimr::skim(wide_data$`Case detection rate (%)`)
+summary_cdr <- skimr::skim(otherburden$`Case detection rate (%)`)
 print(summary_cdr)
 cdrsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "cdrsummarytable.rds")
 saveRDS(summary_cdr, file = cdrsummarytable_file)
 
 #drug resistance summary and save to file location
-summary_drug <- skimr::skim(wide_data$`People with MDR/RR-TB (%)`)
+summary_drug <- skimr::skim(otherburden$`People with MDR/RR-TB (%)`)
 print(summary_drug)
 drugsummarytable_file <- here("results", "exploratory_analysis", "summary tables", "drugsummarytable.rds")
 saveRDS(summary_drug, file = drugsummarytable_file)
 
 ## ---- exploratorytables2 --------
 #incidence summary and save to file location
-summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
+summary_inc <- skimr::skim(otherburden$`TB incidence (new infections per 100 000 population)`)
 print(summary_inc)
[…]

### 2023-05-05 code/analysis_code/statistical_analysis.R

@@ -28,7 +28,7 @@ lm_mod <- linear_reg()
 #fit model with high burden countries
 inclm_fit <- lm_mod %>% 
   fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
-      data = highburden) #fit linear model
+      data = hbmodel) #fit linear model
 incfittable <- tidy(inclm_fit) 
 print(incfittable) #produce tidy table of fitted model
 
@@ -38,7 +38,7 @@ saveRDS(incfittable, file = table_file1) #save summary table
 #fit model with full data
 inclm_fitFD <- lm_mod %>% 
   fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
-      data = wide_data) #fit linear model
+      data = obmodel) #fit linear model
 incfittable2 <- tidy(inclm_fitFD) 
 print(incfittable2) #produce tidy table of fitted model
 
@@ -48,7 +48,7 @@ saveRDS(incfittable2, file = table_file2) #save summary table
 #prediction model using subgroup and setting
 predinclm_fit <- lm_mod %>% 
   fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
-      data = highburden) #fit linear model
+      data = hbmodel) #fit linear model
 predincfittable <- tidy(predinclm_fit) 
 print(predincfittable)
 
@@ -79,7 +79,7 @@ saveRDS(predplot, file = here("results", "statistical_analysis", "model fit tabl
 #fit model with high burden countries
 mortlm_fit <- lm_mod %>% 
   fit(`TB mortality (deaths per 100 000 population)` ~ subgroup,
-      data = highburden) #fit linear model
+      data = hbmodel) #fit linear model
 mortfittable <- tidy(mortlm_fit) 
 print(mortfittable) #produce tidy table of fitted model
 
@@ -89,7 +89,7 @@ saveRDS(mortfittable, file = table_file3) #save summary table
 #fit model with full data
 mortlm_fitFD <- lm_mod %>% 
   fit(`TB mortality (deaths per 100 000 population)` ~ subgroup,
-      data = wide_data) #fit linear model
+      data = otherburden) #fit linear model
 mortfittableFD <- tidy(mortlm_fitFD) 
 print(mortfittableFD) #produce tidy table of fitted model
 
[…]

### 2023-05-05 code/processing_code/processingcode.R

@@ -51,6 +51,7 @@ length(unique(rawdata$setting))
            estimate, population)) %>% 
   pivot_wider(names_from = "indicator_name", values_from = "estimate")
 summarywide <- tbl_sum(wide_data)
+length(unique(wide_data$setting))
 data1 = here("results", "processing", "summarywide.rds")
 saveRDS(summarywide, file = data1)
 
@@ -73,14 +74,36 @@ highburden <- wide_data %>%
            "Kyrgyzstan", "Nepal", "Peru", "Republic of Moldova",
            "Somalia", "Tajikistan", "Ukraine",
            "Uzbekistan", "Zimbabwe")) 
-unique(highburden$setting)
+length(unique(highburden$setting))
 summaryhb <- tbl_sum(highburden)
 data2 = here("results", "processing", "summaryhighburden")
 saveRDS(summaryhb, file = data2)
 
+otherburden <- wide_data %>% 
+  filter(!setting %in% c("Brazil", "Central African Republic", "Congo",
+                        "Ethiopia", "Gabon", "Kenya",
+                        "Lesotho", "Liberia", "Namibia",
+                        "Thailand", "Uganda", "United Republic of Tanzania",
+                        "Botswana", "Cameroon", "Eswatini",
+                        "Guinea", "Guinea-Bissau", "Malawi",
+                        "Russian Federation", "Zimbabwe", "China",
+                        "Democratic Republic of the Congo", "India",
+                        "Indonesia", "Mozambique", "Myanmar",
+                        "Nigeria", "Philippines", "South Africa",
+                        "Zambia", "Sierra Leone", "Angola",
+                        "Bangladesh", "Democratic People's Republic of Korea",
+                        "Mongolia", "Pakistan", "Papua New Guinea",
+                        "Viet Nam", "Azerbaijan", "Belarus", "Kazakhstan",
+                        "Kyrgyzstan", "Nepal", "Peru", "Republic of Moldova",
+                        "Somalia", "Tajikistan", "Ukraine",
+                        "Uzbekistan", "Zimbabwe")) 
+length(unique(otherburden$setting))
+summaryob <- tbl_sum(otherburden)
+data3 = here("results", "processing", "summaryotherburden")
+saveRDS(summaryob, file = data3)
 ## ---- cleandata2 --------
 # look at each indicator to check for cleaning requirements
[…]

### 2023-05-05 products/manuscript/Manuscript.qmd
Final project commit

@@
 After processing the data, the `otherburden` and `highburden` datasets were described by exploring each health equity indicators and tuberculosis outcome.
-The table below shows the list of indicators, and the codebook can be found in the README.md file.
-
-```{r}
-#| label: fig-indicators
-#| fig-cap: "Healthy equity indicators"
-readRDS(file = here("results", "processing", "indicators.rds"))
-```
+The codebook with indicator descriptions can be found in the README.md file.
 
